---
title       : Lab 2
description : Using R to Explore Data Sets

--- type:NormalExercise lang:r xp:100 skills:1 key:f8b383874f
## Getting started with data

For this lab, we will use a dataset called `arbuthnot` which contains data on male and female births in London from 1629-1710.

- `arbuthnot` is already in your workspace so if you type `arbuthnot` in your console, the data set will print.

Sometimes the data can be very long so instead of viewing the entire thing, we can type:

- `head(dataset)` to get the first 6 entries or
- `tail(dataset)` to get the last 6 entries. 

If we want to know the variable (column) names in the data set, we can type `names(dataset)` into the console.

If we want to know the dimensions of a data set (ie. how many variables and observations) use the `dim(dataset)` function.

Note: `dataset` is just a placeholder for the name of the data set you want to explore. In this exercise, you need to substitute `arbuthnot` for `dataset`.

*** =instructions
- Use the `head()` function to view the first 6 entries of the `arbuthnot` data set.
- Use the `tail()` function to view the last 6 entries.
- Find the variable names by using the `names()` function.
- Use `dim()` to find the dimensions of `arbuthnot`.
- Click the 'Submit Answer' button and take a look at the R output in the console.


*** =hint
The data set is named `arbuthnot` so substitute `arbuthnot` for `dataset`.

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Use appropriate function to find first 6 entries of arbuthnot

# Use appropriate function to find last 6 entries of arbuthnot

# Use appropriate function to find the variable names of arbuthnot

# Use appropriate function to find the dimension of arbuthnot


```

*** =solution
```{r}
# Use appropriate function to find first 6 entries of arbuthnot
head(arbuthnot)

# Use appropriate function to find last 6 entries of arbuthnot
tail(arbuthnot)

# Use appropriate function to find the variable names of arbuthnot
names(arbuthnot)

# Use appropriate function to find the dimension of arbuthnot
dim(arbuthnot)

```

*** =sct
```{r}

test_function("head", args = "x", incorrect_msg = "Make sure to use `arbuthnot` instead of `dataset` inside the head() function.")


test_function("tail", args = "x", incorrect_msg = "Make sure to use `arbuthnot` instead of `dataset` inside the tail() function.")


test_function("names", args = "x", incorrect_msg = "Make sure to use `arbuthnot` instead of `dataset` inside the names() function.")


test_function("dim", args = "x", incorrect_msg = "Make sure to use `arbuthnot` instead of `dataset` inside the dim() function.")

success_msg("Good work! Now look at the R output, notice how each horizontal row is a different year. We can see from the `dim()` command that there are 82 years with 3 variables.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:795745ad44
## Review accessing variables

Remember, to access specific variables within a data set use the format:

`dataset$variable`.

You might have seen in the previous exercise that `arbuthnot` has a *number of male births* variable named `boys`. If we want to look at this variable by itself, we can type `arbuthnot$boys` into the R console; this will return the number of male births each year. 


*** =instructions
- The arbuthnot data set also has a *number of female births* variable named `girls`. Use `$` to access the `girls` variable from the `arbuthnot` data set.
*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Access the girls variable from arbuthnot


```

*** =solution
```{r}
# Access the girls variable from arbuthnot
arbuthnot$girls
```

*** =sct
```{r}
test_student_typed("arbuthnot$girls", not_typed_msg = "Make sure you follow the format: dataset$variable with the correct data set and variable!")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:f759511088
## Using the summary() function

In this exercise, you will learn how to use the R built-in function `summary()` to obtain statistics of your data set. 

`summary(dataset)` gives a few summary statistics of all variables in dataset


`summary(dataset$variable)` gives a few summary statistics of a specific variable in dataset

The dataset we've been working with `arbuthnot` is already in your workspace.

*** =instructions
- Use the `summary()` function on the `arbuthnot` dataset.

*** =hint
substitute `arbuthnot` for dataset

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Use summary() on arbuthnot to obtain summary statistics of the arbuthnot dataset


```

*** =solution
```{r}
# Use summary() on arbuthnot to obtain summary statistics of the arbuthnot dataset
summary(arbuthnot)

```

*** =sct
```{r}
test_function("summary", args = "object", incorrect_msg = "Follow the format `summary(dataset)` with `arbuthnot` instead of `dataset` to obtain a summary of the entire data set.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:15bd840872
## Working with variables 

We can also use arithmetic operators on variables. 

For example, if I want to add two variables together, I would type `dataset$var1 + dataset$var2`. 

If I want to save this sum as a new variable in the data set, I would assign it by `dataset$newvar <- dataset$var1 + dataset$var2`.

In this exercise, we will be creating a new variable called `propBoys` which calculates the proportion of male births each year.

$\text{proportion of boy births} = \frac{\displaystyle \text{number of boy births}}{\displaystyle \text{number of boy births} + \text{number of girl births} }$

To do this, we will use the syntax:

`dataset$newvar <- dataset$var1 / (dataset$var1 + dataset$var2)`


*** =instructions
- Use the syntax above to calculate the proportion of boy births for each year in the data set. Name this new variable `propBoys`.
- Recall that:
    - the dataset is `arbuthnot`
    - the number of boy births is given by `boys`
    - the number of girl births is given by `girls`
- Use the `head()` function on `arbuthnot` to verify that `propBoys` was added to the data set.
*** =hint
- Follow the formula in the instructions with specified variables and data set.
- Start typing on the right side of `<-`.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Create propBoys variable for arbuthnot
arbuthnot$propBoys <- 

# Use head() on arbuthnot to check if your new variable was added


```

*** =solution
```{r}
# Create propBoys variable for arbuthnot
arbuthnot$propBoys <- arbuthnot$boys / (arbuthnot$boys + arbuthnot$girls)

# Use head() on arbuthnot to check if your new variable was added
head(arbuthnot)
```
*** =sct
```{r}
test_data_frame("arbuthnot", columns = "propBoys", incorrect_msg = "Make sure to follow the format in the lesson with correct data set and variables.")
test_function("head", args = "x")
success_msg("Great. Now look at the output and verify that `propBoys` was added to the data set.")
```



--- type:NormalExercise lang:r xp:100 skills:1 key:59b987e2d7
## Using the sum() function

Another useful built-in function that R has is the `sum()` function, which simply sums all values.

For example, to sum the values of a variable in a data set use the format:

`sum(dataset$variable)`

*** =instructions
Use the `sum()` function to find the total number of male births `boys` in `arbuthnot` over the years in the data set (1629 - 1710).
*** =hint
- To sum over a variable, use the format: `sum(dataset$variable)`.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Find total number of boy births using the sum() function



```

*** =solution
```{r}
# Find total number of boy births using the sum() function
sum(arbuthnot$boys)


```

*** =sct
```{r}
test_function("sum", args = "...", incorrect_msg = "Make sure you use the format: sum(dataset$variable) with the correct data set and variable!")


```



--- type:NormalExercise lang:r xp:100 skills:1 key:938ccd0b92
## Evaluating logical expressions

Recall in Lab 1 when we reviewed the different types of data.

*Logical* data only takes on two values (TRUE or FALSE) where `TRUE` has the value of 1 and `FALSE` has the value of 0.

We can evaluate *logical* expressions with comparison operators like

* greater than `>`
* less than `<`
* equal to `==`
* not equal to `!=`

For example, a logical expression using the *greater than* symbol should follow the format:

`dataset$var1 > dataset$var2`

This command will return a list of `TRUE` and `FALSE` values for each element evaluated.

We can save these values in an object `new_var` by typing:

`dataset$new_var <- dataset$var1 > dataset$var2`
*** =instructions
- Utilize a logical expression to identify which years boy births (`boys`) exceeds girl births (`girls`).  Assign this result to the variable `boys_gt_girls` in the `arbuthnot` data set.
 - Use the sum() function to evaluate the total number of TRUE results in `boys_gt_girls`; that is, the total number of years in which boy births exceed girl births.
*** =hint
Remember, the format for finding the sum of a variable is `sum(dataset$variable)`.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Create my_var to identify years where boy births exceed girl births
arbuthnot$boys_gt_girls <-

# Use the sum() function on boys_gt_girls to find the total number of years in which boy births exceed girl births
# This has been completed for you
sum(arbuthnot$boys_gt_girls)

```

*** =solution
```{r}
# Create a logical expression using the syntax from the lesson 
arbuthnot$boys_gt_girls <- arbuthnot$boys > arbuthnot$girls


# Use the sum() function on my_var to find how many years there were more male births than female births
# This has been completed for you
sum(arbuthnot$boys_gt_girls)

```

*** =sct
```{r}

test_data_frame("arbuthnot", columns = "boys_gt_girls", incorrect_msg = "Follow the format in the lesson with `arbuthnot` instead of 'dataset', `boys_gt_girls` instead of 'new_var', `boys` instead of 'var1', and `girls` instead of 'var2'.")



test_function("sum", args = "...", incorrect_msg = "Remember, the format for finding the sum of a variable is `sum(dataset$variable)`. In this example, the dataset is `arbuthnot` and the variable is `boys_gt_girls`.")

```




--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:c36eae3a47
## Quick check 1

Suppose I have a variable called GPA in my students data set. What do you think will happen if I enter the following expression?

`sum(students$GPA>3.5)`

*** =instructions
- it will return the sum of all of the GPAs that are greater than 3.5
- there will be an error
- it will return the number of students who have a GPA greater than 3.5
- it will return the number of students who have a GPA less than 3.5
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That's incorrect."
msg_success <- "Good answer!"

test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:6bc90b9c68
## Graphing data: the basics

To create a basic plot in R, use the `plot()` function in the format:

`plot(x = dataset$var1 , y = dataset$var2)`

For this exercise, plot `year` on the x-axis and proportion of male births `propBoys` on the y-axis.


*** =instructions
- Use the `plot()` function to create a scatterplot of `propBoys` by `year` from the `arbuthnot` data set.
- After clicking 'Submit Answer', look at the plot in the output. Is there any pattern?


*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
arbuthnot$propBoys <- arbuthnot$boys / (arbuthnot$boys + arbuthnot$girls)

```

*** =sample_code
```{r}
# Create a scatterplot with year on the x-axis and proportion of male births propBoys on the y-axis


```

*** =solution
```{r}
# Create a scatterplot with year on the x-axis and proportion of male births propBoys on the y-axis

plot(x = arbuthnot$year, y = arbuthnot$propBoys)


```

*** =sct
```{r}
test_function("plot", args = c("x", "y"), incorrect_msg = "Make sure you follow the formula in the lesson with the correct data set and variables!")
success_msg("Good work!")

```


--- type:NormalExercise lang:r xp:100 skills:1 key:d0401021d4
## Adding labels to graphs 

You can specify additional arguments in the `plot()` function:

- `xlab = "x"` labels the x-axis "x"
- `ylab = "y"` labels the y-axis "y"
- `main = "title"` gives a title to the entire plot "title"

To add labels to a plot use the format:

`plot(x = dataset$var1 , y = dataset$var2, xlab = "x", ylab = "y", main = "title")`

Currently, the plot of year versus proportion of boys has the default axis names. In this exercise, we will change the axis names and add a title to the graph.

*** =instructions
- Create the plot from the last exercise `plot(x = arbuthnot$year, y = arbuthnot$propBoys)` but this time add axis labels and a title. 

    - The new x-axis label should be `"Year"`, y-axis label should be `"Proportion of Male Births"` and the title should be `"Arbuthnot: Boys by Year"`.
- Click the 'Submit Answer' button and take a look at the new plot. So much Better!

*** =hint
Make sure to wrap the labels in "" because they are words.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
arbuthnot$propBoys <- arbuthnot$boys / (arbuthnot$boys + arbuthnot$girls)

plot(x = arbuthnot$year, y = arbuthnot$propBoys)
```

*** =sample_code
```{r}
# Fill in the blanks to add axis labels and a title to the plot
plot(x = arbuthnot$year , y = arbuthnot$propBoys, xlab = "____" , ylab = "____", main = "____")


```

*** =solution
```{r}
# Fill in the blanks to add axis labels and a title to the plot
plot(x = arbuthnot$year , y = arbuthnot$propBoys, xlab = "Year", ylab = "Proportion of Male Births", main = "Arbuthnot: Boys by Year")


```

*** =sct
```{r}
test_function("plot", args = c("x", "y", "xlab", "ylab", "main"), incorrect_msg = "Use the format in the lesson with the specified lables in the instructions. Make sure you wrap the labels in quotation marks because they are words.")
success_msg("Look how much better the graph looks now that we have labels!")
```


--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:1037bb100a
## LAST EXERCISE OF LAB 2

**This is the last exercice of Lab 2.**

Which one of the following commands would you use to view the first six entries in a data set?
*** =instructions
- tail()
- summary()
- dim()
- head()
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That's incorrect."
msg_success <- "Good answer! You have completed the Lab 2 practice course. STOP HERE."

test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))

```







