---
title       : Lab 4a
description : Distribution of Sample Proportions and Confidence Intervals for Population Proportions


--- type:NormalExercise lang:r xp:100 skills:1 key:1c2ca885bb
## Taking random samples from a variable

In this lab, you are going to learn how to take random samples from a variable using the `sample()` function.

The `sample()` function takes two arguements:

1) variable that you want to sample (`x`)

2) sample size (`size`)

To take a random sample of a variable use the format:

`sample(x = dataset$variable, size = number)`


In this exercise we'll be using a data set called `peanut_allergy` which contains data on children involved in a peanut allergy study. 

- there are 3 variables in this data set: 
    - `had_early_risk`: indicates whether the child had an early risk of developing a peanut allergy
    - `regimen`: the peanut regimen assigned to the child; either 'consume' or 'avoid' peanuts
    - `allergic`: indicates whether the child developed a peanut allergy by the end of the study
- there are 628 subjects 
- `peanut_allergy` is already in your workspace

*We will treat this data set as the entire population of interest.*

*** =instructions
- Take a random sample of 25 from the variable `allergic` in the `peanut_allergy` dataset.
- Take a random sample of 50 from the variable `regimen` in the `peanut_allergy` dataset.
*** =hint
For the 1st instruction, follow the format in the lesson but use `x = peanut_allergy$allergic` and `size = 25`.

For the 2nd instruction, follow the format in the lesson but use `x = peanut_allergy$regimen` and `size = 50`.
*** =pre_exercise_code
```{r}
source("https://www.openintro.org/stat/data/peanut_allergy.R")


```

*** =sample_code
```{r}
# take a random sample of 25 from the allergic variable


# take a random sample of 50 from the regimen variable


```

*** =solution
```{r}
# take a random sample of 25 from the allergic variable
sample(x = peanut_allergy$allergic, size = 25)

# take a random sample of 50 from the regimen variable
sample(x = peanut_allergy$regimen, size = 50)

```

*** =sct
```{r}
test_function("sample", args = c("x", "size"), index = 1, incorrect_msg = "Follow the format in the lesson with the specified dataset (peanut_allergy), variable (allergic), and number (25)!")
test_function("sample", args = c("x", "size"), index = 2, incorrect_msg = "Follow the format in the lesson with the specified dataset (peanut_allergy), variable (regimen), and number (50)!")

```

--- type:NormalExercise lang:r xp:100 skills:1 key:190f70cfa4
## Random samples cont...

In this exercise, we are going to take two random samples and assign them to object names using the assignment notation `<-`. 


Note: Remember that after we create an object (i.e. sample1 and sample2) we need to type the object name to view it *(I have already typed this for you in the R Script)*.
*** =instructions
- Take a random sample of 20 from the variable `regimen` in the `peanut_allergy` dataset, assign this to `sample1`.
- Take another random sample of 20 from the same variable, but assign this sample to `sample2`.
- Type the command `sample1 == sample2` in your script and click the 'Submit Answer' button.
*** =hint
Make sure you follow the format: `sample(x = dataset$variable, size = number)` with specified dataset, variable and number.
*** =pre_exercise_code
```{r}
source("https://www.openintro.org/stat/data/peanut_allergy.R")

```

*** =sample_code
```{r}
# take a random sample of 20 from the variable regimen
sample1 <- 

sample1

# take another random sample of 20 from the variable regimen
sample2 <- 

sample2

# Type the command 'sample1 == sample2' below to see if the samples are equal


```

*** =solution
```{r}
# take a random sample of 20 from the variable regimen
sample1 <- sample(peanut_allergy$regimen, 20)


sample1

# take another random sample of 20 from the variable regimen
sample2 <- sample(peanut_allergy$regimen, 20)


sample2

# Type the command 'sample1 == sample2' below to see if the samples are equal
sample1 == sample2

```

*** =sct
```{r}
test_object("sample1", incorrect_msg = "Make sure you follow the format: sample(x = dataset$variable, size = number) with specified dataset, variable and number.")

test_object("sample2", incorrect_msg = "Make sure you follow the format: sample(x = dataset$variable, size = number) with specified dataset, variable and number.")

test_output_contains("sample1 == sample2", incorrect_msg = "Make sure you type 'sample1 == sample2' in your script before submitting your answer.")

success_msg("Nice job! Look at the R output in the bottom right panel. Do the contents of sample1 and sample2 match up? We can easily answer this by examining the TRUE's and FALSE's from the 'sample1 == sample2' command; if there are FALSEs then we know our samples are not the same.")

```


--- type:NormalExercise lang:r xp:100 skills:1 key:2217632b88
## Using the set.seed() function

You probably noticed that two samples you took (of the same variable and size) were not the same. 

This is because `sample1` and `sample2` are *random* samples; so each time R takes a random sample of 20 from `regimen`, the values will be different.

Sometimes we want to have a *consistent* random sample. To do this, we use the `set.seed(number)` function. If you type this **before** creating your random sample, your sample will be the ***same*** random sample every time you Knit the document.

Note: Any number can go inside the `set.seed()` function.
*** =instructions
- Type `set.seed(1234)` in the R Script.
- Now, use the `sample()` function to get a random sample of 20 from the `allergic` variable in the `peanut_allergy` dataset.
*** =hint
- You can copy and paste `set.seed(1234)` or type `set.seed(__)` with any number in the parenthesis.

- To take a random sample of a variable use the format:

`sample(dataset$variable, number)`

*** =pre_exercise_code
```{r}

source("https://www.openintro.org/stat/data/peanut_allergy.R")

```

*** =sample_code
```{r}
# Type set.seed(1234)

# take a random sample of 20 from the allergic variable 



```

*** =solution
```{r}
# Type set.seed(1234)
set.seed(1234)

# take a random sample of 20 from the allergic variable 
sample(peanut_allergy$allergic, 20)
```

*** =sct
```{r}
test_function("set.seed", args = "seed", eval = NA, incorrect_msg = "Copy and paste the set.seed() function from the first instruction. You can can change the number inside the function if you want; I arbitrarily chose '1234' but any whole number will work.")

test_function("sample", args = c("x", "size"), incorrect_msg = "Remember, to take a random sample from a variable use the format: sample(dataset$variable, number) with the specified parameters.")

```


--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:5b1897d9be
## Quick check 1

Suppose I want to take a sample of size 35 from the `regimen` variable in the `peanut_allergy` data set, and store those results in an object called mysample.  Which of the following commands is correct?

*** =instructions
- mysample <- sample(peanut_allergy$regimen, 35)
- mysample <- sample(regimen, 35)
- mysample <- sample(35, regimen)
- sample(peanut_allergy$regimen, 35) <- mysample
- sample(mysample,35) <- regimen

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. Remember, the format to take a sample is sample(dataset$variable, number)."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad, msg_bad))


```

--- type:NormalExercise lang:r xp:100 skills:1 key:b86b269872
## Population proportions

For this exercise, you are going to find the *population proportion* of children who developed an allergic reation to peanuts.

**Remember, we will treat this data set as the entire population of interest.**



*** =instructions
- Use the `table()` function to get an appropriate summary of the `allergic` variable in the `peanut_allergy` data set.
- Type the command `addmargins(table(peanut_allergy$allergic))` to view the table with row/column totals.
- Click the 'Submit Answer' button and look at the R output.

*** =hint
The correct format to summarize a categorical variable is:

`table(dataset$variable)`.

For the last instruction, divide the number of children who were allergic by the total number of children.
*** =pre_exercise_code
```{r}
source("https://www.openintro.org/stat/data/peanut_allergy.R")
```

*** =sample_code
```{r}
# Create a table of the `allergic` variable in the `peanut_allergy` dataset

# Copy and paste the command in the second instruction to view row/column totals of the table



```

*** =solution
```{r}
# Create a table of the `allergic` variable in the `peanut_allergy` dataset
table(peanut_allergy$allergic)

# Copy and paste the command in the second instruction to view row/column totals of the table
addmargins(table(peanut_allergy$allergic))


```

*** =sct
```{r}
test_function("table", args = "...", incorrect_msg = "Remember, the correct format for the table() function is: table(dataset$variable). Create a table with this format using the specified data set and variable.")

success_msg("Nice! Now divide the number of children who were allergic by the total number of children in the data set. This will give you the 'population' proportion of children with a peanut allergy.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:38ce97a2e8
## Sample proportions

Most of the time we don't know what the *population proportion* is, but we can estimate this value with a random sample. 

In this exercise, you will create a random sample of 50 from the `allergic` variable in the dataset `peanut_allergy` and then compare the *sample proportion* of allergic children to the *population proportion* you found in the last exercise.
*** =instructions
- Use the `sample()` function to create a sample of 50 from the `allergic` variable, assign this to `my_sample`.
- Use the `table()` function on `my_sample`.
- Click 'Submit Answer' and look at the R output to find how many children were allergic in `my_sample`.
*** =hint
For the first instruction, use the format `sample(x = dataset$variable, size = number)`.

For the last instruction, divide the number of children who were allergic by the total number of children in the sample.

*** =pre_exercise_code
```{r}
source("https://www.openintro.org/stat/data/peanut_allergy.R")

```

*** =sample_code
```{r}
# Use the format sample(x = dataset$variable, size = number) to get a sample of 50 from the `allergic` variable
my_sample <- 

# Type table(my_sample) to see how many children were allergic in the sample



```

*** =solution
```{r}
# Use the format sample(x = dataset$variable, size = number) to get a sample of 50 from the `allergic` variable
my_sample <- sample(peanut_allergy$allergic, 50)

# Type table(my_sample) to see how many children were allergic
table(my_sample)



```

*** =sct
```{r}
test_object("my_sample", incorrect_msg = "Make sure you follow the format: sample(x = dataset$variable, size = number) with specified dataset, variable and number, then assign it to my_sample.")

test_function("table", args = "...", incorrect_msg = "Type table(my_sample) to get a numerical summary of my_sample.")

success_msg("Nice! Now divide the number of children in your sample (my_sample) who were allergic by the total number of children in your sample. This will give you the 'sample' proportion of children with a peanut allergy. How does this compare to the population proportion?")
```




--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:0915d27a1d
## Quick check 2

Suppose that out of a sample of 50 children, 9 developed an allergy to peanuts. How does this sample proportion compare to the population proportion?

*Remember, 64 children had a peanut allergy out of the 'population' of 628 children.*
*** =instructions
- The sample proportion and population proportion are equal.
- The sample proportion is greater than the population proportion.
- The sample proportion is less than the population proportion.

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly!"
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_success, msg_bad))

```

--- type:NormalExercise lang:r xp:100 skills:1 key:0e23913548
## Confidence intervals for a population proportion

Assume that the population proportion of children with peanut allergies is unknown so we obtained a random sample. 

In this exercise, we are going to use R as a calculator to find a 95% confidence interval for the true proportion of children who are allergic to peanuts *(population proportion)*.

Below is the formula to construct a confidence interval for a population proportion.

$\hat{y} \pm z^*\times \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$

*** =instructions
Suppose I sampled 50 of the children from `peanut_allergy` and found that 0.15 reported *yes* to having a peanut allergy. Use R as a calculator to find a 95% confidence interval for the population proportion based on this sample data (Use the formula in the lesson).

Note that the $z^*$ critical value is 1.96 for a 95% confidence interval.

*Remember, to square a number use the format: `sqrt(number)`*
*** =hint
- Find 'upper bound' and 'lower bound' seperately.
- Use the formula in the lesson with $\hat{p}$ = 0.15, and n = 50.
- When multiplying two things together use an asterisk $\ast$. For example if we want to multiply 4 and 8, type `4*8`; typing `4(8)` will result in an error.
*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}
#Lower confidence bound

#Upper confidence bound



```

*** =solution
```{r}
#Lower confidence bound
0.15 - 1.96 * sqrt(.15*.85/50)

#Upper confidence bound
0.15 + 1.96 * sqrt(.15*.85/50)

```

*** =sct
```{r}
test_output_contains("0.05102485", incorrect_msg = "Follow the order of operations from your high school algebra class 'PEMDAS' and remember, when you want to square something in R, use the sqrt() function. For the lower confidence bound, subtract the margin of error from the sample proportion.")

test_output_contains("0.2489751", incorrect_msg = "Follow the order of operations from your high school algebra class 'PEMDAS' and remember, when you want to square something in R, use the sqrt() function. For the upper confidence bound, add the margin of error to the sample proportion.")


```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:a4477e43a4
## Quick check 3

**This is the last exercise of Lab 4a**

Which of the following are valid implementations of the `set.seed()` command?  


*** =instructions
- set.seed("abc")
- set.seed("1234")
- set.seed(xyz)
- set.seed(5678)
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. Remember, any number can go in the set.seed() function but there is no quotations."
msg_success <- "Exactly! You finished Lab 4a practice course."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))


```
