---
title       : Lab 5
description : One-sample t-test


--- type:NormalExercise lang:r xp:100 skills:1 key:fe45e251b5
## One sample t-test

The one sample t-test is used when we want to test a specific value for the population mean based on our sample data.

To perform a one sample t-test, we'll be using a data set called `HELPrct` which contains data on adult inpatients who were recruited from a detoxification unit. 

- there are 27 variables in this data set
- there are 453 subjects 
- `HELPrct` is already in your workspace

`HELPrct` contains a variable called `i1` (i-one) which reports the typical number of alcoholic drinks consumed ***per day*** (within the last month). 

If you want more information about this data set type: `?HELPrct` into the R console.
*** =instructions
- Type `HELPrct$i1` in your script to view the typical number of alcoholic drinks the patients consumed per day before coming to the detox facility.
- Press the 'Submit Answer' button and take a look at the R output. 

*** =hint

*** =pre_exercise_code
```{r}

library(mosaicData)

```

*** =sample_code
```{r}
# type the code in the instructions and press 'Submit Answer'


```

*** =solution
```{r}
# type the code in the instructions and press 'Submit Answer'
HELPrct$i1


```

*** =sct
```{r}
test_student_typed("HELPrct$i1", not_typed_msg = "Remember to type HELPrct$i1 in your script.")

success_msg("Good. Now scroll through the R output to see the average number of alcoholic drinks per day for each subject. Do these numbers suprise you?") 
```
--- type:NormalExercise lang:r xp:100 skills:1 key:9c0a9256c3
## Assumptions of the t-test

According to the *Dietary Guidelines for Americans*, healthy alcohol consumption is defined by having about 2 drinks per day. Suppose we want to test if the average number of alcoholic drinks is 2 based on our sample.


Before conducting a one sample t-test, we need to make sure that our data meets these assumptions:

* the observations are independent
* the variable is normally distributed (or large sample size)

It is reasonable to assume that the number of drinks consumed by individuals pre-treatment are not related to each other, so the observations are independent.

In this exercise, you will be checking the normality assumption for the `i1` variable in `HELPrct`.

*The second condition is met if a histogram of the variable looks approximately symmetric and bell-shaped OR if sample size is greater than 30.*


*** =instructions
- Type `library(mosaic)` to load the *mosaic* package into your workspace.
- Use the `favstats()` function to examine the distribution of `i1` from the `HELPrct` data set.
- Create a histogram of `i1` from the `HELPrct` data set using the `hist()` function.
*** =hint
To obtain summary statistics for a quantitative variable, use the format `favstats(dataset$variable)`.

To create a histogram, use `hist(dataset$variable)`.



*** =pre_exercise_code
```{r}
library(mosaicData)
library(mosaic)
```

*** =sample_code
```{r}
# Load the mosaic package 


# Find the summary statistics of 'i1' in the HELPrct data set using the favstats() function

# Check for normality using the hist() function


```

*** =solution
```{r}
# Load the mosaic package
library(mosaic)

# Find the summary statistics of 'i1' using the favstats() function
favstats(HELPrct$i1)

# Check for normality using the hist() function
hist(HELPrct$i1)

```



*** =sct
```{r}
test_student_typed("library(mosaic)",  not_typed_msg =  "Type library(mosaic) to load the mosaic package.")


test_function("favstats", args = "x", incorrect_msg = "Make sure you follow the format favstats(dataset$variable) with the HELPrct dataset and 'i1' variable.")

test_function("hist", args = "x", incorrect_msg = "Make sure you follow the format hist(dataset$variable) with the HELPrct dataset and 'i1' variable.")

success_msg("Nice!")
```





--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:af0f4164c6
## Quick check 1
The *Plots* panel contains the histogram of the typical number of drinks (`i1`) from the last exercise. 

The `favstats()` results from the last exercise are stored in an object called `results`; in your R console type `results` to view these.

Question: Is the second condition (normality) satisfied for our data?

*Remember, the second condition is met if a histogram of the variable looks approximately symmetric and bell-shaped OR if sample size is greater than 30.*
*** =instructions
- No, the distribution is right-skewed and the sample size is less than 30.
- Yes, the distribution is left-skewed but the sample size is greater than 30.
- No, the distribution is normal but the sample size is less than 30.
- Yes, the distribution is right-skewed but the sample size is greater than 30.
*** =hint
Since the tail of the distribution is on the right side, it is right-skewed. Type `results` to view the sample size.
*** =pre_exercise_code
```{r}
library(mosaicData)
library(mosaic)
hist(HELPrct$i1)
results <- favstats(HELPrct$i1)
```

*** =sct
```{r}
msg_bad <- "That is not correct. The tail of the distribution is on the right side so it is right-skewed BUT the sample size is 453."
msg_success <- "Exactly! Now that we know the conditions are satisfied, we can conduct a one sample t-test."
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```
--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:b8967b17fb
## Quick check 2

Which function in R allows you to simultaneously obtain the mean and standard deviation of quantitative variable?


*** =instructions
- summary()
- head()
- favstats()
- mosaic()

*** =hint
The summary function gives you the mean but not the standard deviation. Remember, `mosaic` is a package not a function.
*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. The summary function gives you the mean but not the standard deviation. Remember, `mosaic` is a package not a function."
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:52ad764c46
## Using the t.test() function

Suppose we want to:

* perform a t-test to determine if the population mean number of drinks per day differs from 2, and, 
* build a 95% confidence interval to estimate the true mean number of drinks. 

We can do this in one step with the `t.test()` function.

The default settings are to test a 2-sided alternative hypothesis and to calculate a 95% confidence interval.

So, to carry out a one sample t-test for a 2-sided alternative and a 95% confidence interval, use the format:

`t.test(x = dataset$variable, mu = null.hyp)`



Our hypotheses are:

$H_0$: $\mu \= 2$

$H_a$: $\mu \neq 2$,  where $\mu$ is the population average number of alcoholic drinks per day 



*Note: null.hyp is a placeholder for our null hypothesis value.*


*** =instructions
- Use the `t.test()` function to test the hypothesis (stated above) and to create a 95% confidence interval to estimate the population mean number of drinks per day `i1` in the `HELPrct` data set.

- click 'Submit Answer' and look at the R output.

*** =hint
- follow format from lesson with `HELPrct$i1` instead of dataset$variable and  `2` instead of null_hyp
*** =pre_exercise_code
```{r}
library(mosaicData)


```

*** =sample_code
```{r}
# Perform a t-test using the t.test() function 



```

*** =solution
```{r}
# Perform a t-test using the t.test() function 
t.test(x = HELPrct$i1, mu = 2)

```

*** =sct
```{r}
test_function("t.test", args = c("x", "mu"), incorrect_msg = "Make sure you substitute the correct dataset, variable and null hypothesis! In this case we're using the 'HELPrct' data set, 'i1' variable, and null hypothesis equal to 2.")
success_msg("Great job! Now look at the R output and find the p-value for the test and the 95% confidence interval for the mean number of alcoholic drinks per day. How would you conclude this test?")

```




--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:b96cc040de
## Quick check 3

The `t.test` results from the last exercise are stored in an object called `test_results`; in your R console type `test_results` to view these.

At the $\alpha = 0.05$ level of significance, how would we conclude this t-test?

*** =instructions
- Reject the null hypothesis and conclude that the true mean number of alcoholic drinks consumed per day is significantly greater than 2.
- Fail to reject the null hypothesis and conclude that the true mean number of alcoholic drinks consumed per day is significantly greater than 2.
- Reject the null hypothesis and conclude that the true mean number of alcoholic drinks consumed per day is significantly less than 2.
- Fail to reject the null hypothesis and conclude that the true mean number of alcoholic drinks consumed per day is not significantly different from 2.

*** =hint
Type `test_results` and find the p-value for this test. If it is less than 0.05 then we would REJECT the null hypothesis. 
*** =pre_exercise_code
```{r}
library(mosaicData)
test_results <- t.test(x = HELPrct$i1, mu = 2)
```

*** =sct
```{r}
msg_bad <- "That is not correct. Since the p-value is almost zero, we would reject the null hypothesis."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))

```
--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:d122443eee
## Quick check 4


The default settings for a t.test function (eg. `t.test(x = HELPrct$i1, mu = 2`) ) are to  perform a _________ alternative hypothesis and to calculate a ______ confidence interval.

*** =instructions
- two-sided, 90%
- two-sided, 95%
- one-sided, 95%
- one-sided, 90%
*** =hint
In the last exercise we used the `t.test()` default values.
*** =pre_exercise_code
```{r}

```

*** =sct
```{r}

msg_bad <- "That is not correct"
msg_success <- "Exactly!"
test_mc(correct = 2, feedback_msgs = c(msg_bad, msg_success, msg_bad, msg_bad))

```

--- type:NormalExercise lang:r xp:100 skills:1 key:eeccda0836
## One-sided hypothesis tests

Sometimes we have a one-sided alternative hypothesis, for example: 

$H_0$: $\mu \= 2$

$H_a$: $\mu \> 2$,  
where $\mu$ is the population average number of alcoholic drinks per day 


The `t.test()` function can still do this, but we have to specify an additional argument.

To specify an alternative hypothesis, use the format:

`t.test(x = dataset$variable, mu = null_hyp, alternative = "____")`


If our alternative hypothesis is *less than*, use:

`alternative = "less"`

If our alternative hypothesis is *greater than*, use:

`alternative = "greater"`

If our alternative hypothesis is *two-sided*, we can use:

`alternative = "two.sided"`

* The `t.test()` default setting are to perform a two sided test so you don't have to specify `alternative = "two.sided"` for a two sided test.


*** =instructions
Use the `t.test()` function to carry out a one sample t-test on the `HELPrct` data set to determine if the population mean number of drinks per day is **greater** than 2.


*** =hint
- The code is the same as the last time with the added argument: `alternative = "greater"`.
*** =pre_exercise_code
```{r}
library(mosaicData)


```

*** =sample_code
```{r}
# use the t.test() function to determine if the 'i1' is significantly greater than 2 on average


```

*** =solution
```{r}
# use the t.test() function to determine if the 'i1' is significantly greater than 2 on average

t.test(x = HELPrct$i1, mu = 2, alternative = "greater")

```

*** =sct
```{r}
test_function("t.test", args = c("x", "alternative", "mu"), incorrect_msg = "Follow the format in the lesson. Everything is the same as the last exercise except for the added argument- the alternative argument should be equal to 'greater'.")


```



--- type:NormalExercise lang:r xp:100 skills:1 key:198408a028
## Specifying confidence levels

We can also change the confidence level by adding the argument: `conf.level = 0.__` where you write the level as a decimal between 0 and 1.

For example, if I want to test whether the population mean number of drinks per day differs from 2 AND build a **99%** confidence interval for the mean, I would type:

`t.test(x = HELPrct$i1, mu = 2, conf.level = 0.99)`

Note: The default confidence level is 0.95 so for a 95% confidence interval we do not have to specify `conf.level = 0.95`. 

*** =instructions
Use the `t.test()` function to carry out a one sample t-test to determine if the population mean number of drinks per day `i1` *differs* from 2 on average and use a **90% confidence interval** to estimate the population mean.


Our hypotheses are:

$H_0$: $\mu \= 2$

$H_a$: $\mu \neq 2$,  where $\mu$ is the population average number of alcoholic drinks per day 

*** =hint
Make sure you specify the confidence level as a *proportion*. For example, to do a 80% confidence interval add `conf.level = 0.80`. 


This exercise is asking you to create a 90% confidence interval.
*** =pre_exercise_code
```{r}
library(mosaicData)
```


*** =sample_code
```{r}
# Carry out a t-test and find a 90% confidence interval


```

*** =solution
```{r}
# Carry out a t-test and find a 90% confidence interval
t.test(x = HELPrct$i1, mu = 2, conf.level = 0.90)
```

*** =sct
```{r}
test_function("t.test", args = c("x", "mu", "conf.level"), incorrect_msg = "Have you set the conf.level equal to 0.90? Make sure you follow the t.test() format with 'HELPrct' data set, 'i1' variable, mu equal to 2, and conf.level = 0.90.")

success_msg("Good job. If you scroll through the R output you'll notice that the p-value is almost zero!")

```




--- type:NormalExercise lang:r xp:100 skills:1 key:5d2e26a906
## t.test() practice

Conclusion from the last exercise:

At  $\alpha$ =  0.05, we would REJECT the null hypothesis and conclude that the true mean number of alcoholic drinks consumed per day is significantly **more** than 2. This makes sense because the data was taken from patients at a rehab facility. 

Here is a different hypothesis for this population:

$H_0$: $\mu \= 15$

$H_a$: $\mu \> 15$,  where $\mu$ is the population average number of alcoholic drinks per day 

Note: to use the `t.test()` function use format:

`t.test(x = dataset$variable, mu = null.hyp, alternative = "__", conf.level = 0.__)`



*** =instructions
- Use the `t.test()` function to carry out a one sample t-test for the above hypothesis and construct a 99% confidence interval. 
    - this is a one-sided test so you will need to need to specify `alternative="___"`
    - this is **not** a 95% confidence interval, you need to specify `conf.level = 0.__`

*Remember, we are using the `i1` variable from the `HELPrct` data set.*
*** =hint
- This is a one-sided **greater** than test, so `alternative="greater"` should be specified in the `t.test()` function. 

- Make sure you specify the confidence level as a *proportion*. For example, to do a 80% confidence interval add `conf.level = 0.80`. 
*** =pre_exercise_code
```{r}
library(mosaicData)


```

*** =sample_code
```{r}
# Use t.test to perform a t test on the specified hypothesis



```

*** =solution
```{r}
# Use t.test to perform a t test on the specified hypothesis 
t.test(x = HELPrct$i1, mu = 15, alternative = "greater", conf.level = 0.99)

```

*** =sct
```{r}
test_function("t.test", args = c("x", "mu", "alternative", "conf.level"), incorrect_msg = "Have you set the conf.level equal to 0.99 and alternative equal to 'greater'? Make sure you follow the format in the lesson with the specified arguements.")

success_msg("Great!")
```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:878a231ebf
## Quick check 5

What is the correct R command to construct a 90% confidence interval on the following hypothesis: 

$H_0$: $\mu \= 5$

$H_a$: $\mu \< 5$
*** =instructions
- t.test(x = dataset$variable, alternative = "less", conf.level = 0.90)
- t.test(x = dataset$variable, mu = 5, alternative = "greater")
- t.test(x = dataset$variable, mu = 5, alternative = "less", conf.level = 0.90)
- t.test(x = dataset$variable, mu = 5, alternative = "less", conf.level = 90)

*** =hint
Remember, the default settings are to perform a 2-sided alternative hypothesis and to calculate a 95% confidence interval.

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. Select the option that tests for a one-sided 'greater than' hypothesis and confidence level of 0.90"
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:4579e2c495
## Quick check 6

**This is the last exercise of Lab 5**


Which argument should we specify in the t.test function to do a one sided alternative hypothesis in a one sample t-test?

*** =instructions
- conf.level
- lower.tail
- upper.tail
- alternative 
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct"
msg_success <- "Exactly! You finished Lab 5 practice course."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))

```

