---
title       : Lab 1
description : Introduction to R and RStudio




--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:9707072219
## Capabilities of R



What can R be used for?

*** =instructions
- Graphing data
- Analyzing data
- Calculations
- All of the above

*** =hint
R software can be used for all of these things.

*** =pre_exercise_code
```{r}
# The pre exercise code runs code to initialize the user's workspace.
# You can use it to load packages, initialize datasets and draw a plot in the viewer

```

*** =sct
```{r}
# SCT written with testwhat: https://github.com/datacamp/testwhat/wiki

msg_bad <- "That is not correct"
msg_success <- "Exactly! R can do all of these things."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))
```


--- type:NormalExercise lang:r xp:100 skills:1 key:8902293ef6
## How to use DataCamp

The upper right portion of your screen is the script editor: this is where you will be typing your R code to answer the questions. When you press the 'Submit Answer' button, every line of code is interpreted and executed by R and you will get a message that says whether or not your code was correct. 

Below the script is the R Console: you can use this to execute R commands and test your R code for correctness before you submit it. 

The # symbol lets you comment on your code without R processing it.

*** =instructions 
This exercise is only meant to help you get set up with DataCamp so there is no coding task.
Just click 'Submit Answer' to continue to the next exercise.
*** =hint

*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}
# This is were you will be writing your code



```

*** =solution
```{r}
# This is were you will be writing your code



```

*** =sct
```{r}

```
--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:77ff5d2cfd
## Getting started with R


The best way to learn R programming is to actually practice typing yourself. 
So, lets start coding!

Are you excited to dive into the exhilarating world of R Programming?!

*** =instructions
- Yes
- No

*** =hint
You should be very excited!
*** =pre_exercise_code
```{r}

```


*** =sct
```{r}
msg_bad <- "Well, hopefully it will grow on you!"
msg_success <- "Good answer!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad))
```

--- type:NormalExercise lang:r xp:100 skills:1 key:dbdb4a1084
## Using R as a calculator


In this exercise, we'll see how R can be used to perform simple arithmetic tasks.

Some arithmetic operators you can use in R code include:

Addition: `+`

Subtraction: `-` 

Division: `/` 

Multiplication: `*` 

Exponentiation: `^`

For example, if I want to cube '4', I would type `4^3` into the console.

Note: When multiplying two things together use an asterisk $\ast$. For example if we want to multiply 4 and 8, type `4*8`; typing `4(8)` will result in an error.

*** =instructions
- In the script, type `4^3` to calculate four to the power of three.
- Click the 'Submit Answer' Button and take a look at the R output in the console (bottom right panel).
*** =hint
- Type your code in the upper right box


*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}

# Example of Addition
5 + 10

# Example of division and subtraction
(10 - 5) / 2

# Use R as a calculator to find four to the power of three.




```


*** =solution
```{r}

# Example of Addition
5 + 10

# Example of division and subtraction
(10 - 5) / 2

# Use R as a calculator to find four to the power of three
4^3



```

*** =sct
```{r}
test_output_contains("15", incorrect_msg = "Do not remove the line of R code that calculates the sum of 5 and 10. Instead, just do the last question")


test_output_contains("2.5", incorrect_msg = "Do not remove the line of R code that calculates (10 - 5) / 2. Instead, just do the last question")

test_output_contains("64", incorrect_msg = "Make sure to add a line of R code, that calculates 4 to the power of 3. Do not start the line with a `#`, otherwise, your R code will not be executed!")

success_msg("Awesome! See how the console shows the result of the R code you submitted? Now that you're familiar with the interface, let's get down to R business!")
```



--- type:NormalExercise lang:r xp:100 skills:1 key:b19e862406
## Using the sqrt() function

Like any other calculator, R has the ability to take the square root of non-negative numbers.

We will be calculating the square root a lot in this class so knowing this R function is really handy.

If I want to take the square root of 9, I would type `sqrt(9)` and it would return 3.

*** =instructions
- In the script, use the `sqrt()` function to find the square root of 9659.
- In the script, use the `sqrt()` function and arithmetic operators (from last exercise) to evaluate the expression:
$\frac{0.6 - 0.5}{\sqrt{0.25 / 1000}}$


*** =hint
R uses PEMDAS order of operations like you would use in any math class.

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Find the square root of 9659 using sqrt()

# Evaluate the expression in the plots panel 

```

*** =solution
```{r}
# Find the square root of 9659 using sqrt()
sqrt(9659)

# Evaluate the expression in the plots panel 
(0.6-0.5) / sqrt(0.25/1000)
```

*** =sct
```{r}
test_function("sqrt", args = "x")

test_output_contains("(0.6-0.5) / sqrt(0.25/1000)", incorrect_msg = "That is incorrect. Type (0.6-0.5) / sqrt(0.25/1000) to evaluate the expression in the Plots panel.")

success_msg("Good work!")
```

--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:6f441f4fb9
## Basic data types in R

The four basic data types we will be working with in this course are:

Numeric: numbers with decimals like `3.1` or `22.0`

Integers: natural numbers like `-9` and `4`, integers are also numeric data.

Logical: boolean values (`TRUE` or `FALSE`)

Character: texts and words ("this is character data")


Note: all character data should be wrapped in "".

If I typed `70.0 + 28.9` into the console, the value that R would return would be what type of data?

*** =instructions
- numeric
- character
- logical
- integer

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct, '70.0 + 28.9' does not result in a whole number so the data type must be 'numeric'."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))
```


--- type:NormalExercise lang:r xp:100 skills:1 key:8c5519a279
## Variable assignment

'Variable' is one of the most important concepts in statistics. In R, we can use variables to store values or objects so that we can easily access them in the future.

A simple example of variable assignment is `x <- 5` which assigns a value 5 to a variable named 'x'. We can print the contents of x by typing `x` into the R console.


*** =instructions
- Assign the value 100 to a variable named `my_variable`.
- Type `my_variable` into your R script to print the contents.
- Click the 'Submit Answer' button and look through the R output.

*** =hint
Type '100' on the right side of the assignment command `<-` to assign 100 to `my_variable`.
*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Assign the value 100 to a variable named my_variable

my_variable <- 


# Print the value of my_variable


```

*** =solution
```{r}
# Assign the value 100 to a variable named my_variable

my_variable <- 100


# Print the value of my_variable
my_variable

```

*** =sct
```{r}
test_object("my_variable", incorrect_msg = "Make sure you type '100' on the right side of the assignment command `<-` to assign 100 to my_variable.")

test_output_contains("my_variable", incorrect_msg = "Remember to type 'my_variable' to print out the contents of my_variable") 

success_msg("Good! Notice that the value 100 appears when R prints `my_variable`.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:21fd16b05f
## Accessing specific variables using $
R has a built-in dataset called `women` which contains 2 variables (`height` and `weight`) of 15 women. 

This data set is already in your workspace so if you type  `women` into your R console, the data set will come up in the output.


In this exercise you will be learning how to access specific variables within a data set by using the format:

`dataset$variable`.

For example, if you want to view the `height` variable by itself, type `women$height` into the R console.

*** =instructions
Use the format in the lesson to access the `weight` variable in the `women` data set.
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Access weight from women like we did for the height variable


```

*** =solution
```{r}
# Access weight from women like we did for the height variable
women$weight
```

*** =sct
```{r}
test_student_typed("women$weight", not_typed_msg = "Make sure you follow the format: dataset$variable with the correct data set and variable!")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:766b6a5af5
## Graphing data in R

In this exercise, you will practice using simple graphical functions in R to visualize data.

In it's most basic form, the `plot()` function will generate a scatterplot of the x and y values that have been passed through it. 

The format to make a basic scatterplot is:

`plot(x = dataset$x.var , y = dataset$y.var)`

For this exercise, we are going to visualize the relationship between the two variables in `women`:

x.var = `height` 

y.var = `weight`

*** =instructions
- Follow the `plot()` function format from the lesson to create a scatterplot of `height` by `weight` from the `women` data set.
- After clicking 'Submit Answer', look at the plot in the output. Is there any pattern?
*** =hint
Follow the format in the lesson with `women` instead of `dataset`, `height` instead of `var1` and `weight` instead of `var2`.
*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Use plot() to create scatterplot of height by weight


```

*** =solution
```{r}
# Use plot() to create scatterplot of height by weight

plot(x = women$height, y = women$weight)
```

*** =sct
```{r}
test_function("plot", args = c("x", "y"), incorrect_msg = "In your R script, follow the format: plot(x = dataset$var1 , y = dataset$var2) with the specified arguements. In this case, use `women` instead of `dataset`, `height` instead of `var1` and `weight` instead of `var2`.")
success_msg("Good work! From the plot we can see that as height increases, weight decreases.")

```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:dff314deba
## Quick check 
**This is the last exercise of Lab 1**

What is the correct code to access a variable in a data set?

*** =instructions
- dataset#variable.name
- dataset&variable.name
- dataset=variable.name
- dataset$variable.name
- dataset%variable.name
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. We use a dollar sign '$' to access variables."
msg_success <- "Exactly! You finished Lab 1 practice course."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success, msg_bad))


```
