---
title       : Lab 7
description : Inference for Categorical Data



--- type:NormalExercise lang:r xp:100 skills:1 key:616c48f5cf
## Exploring a single categorical variable

In this course, we will use the one sample *z*-test to test whether a population **proportion** differs from a specific value.

To perform a one sample z-test, we'll be using a data set called `nc` which contains data on the North Carolina birth records from 2004. 

- there are 13 variables in this data set 
- there are 1,000 subjects 
- `nc` is already in your workspace

`nc` contains a categorical variable called `lowbirthweight` with 2 levels:

- *low* (low birthweight)
- *not low* (not low birthweight)

Suppose we are interested in the *proportion* of low birthweights.

Let's begin by exploring the `lowbirthweight` variable in the `nc` data set.

*** =instructions
- Use the `table()` function to obtain a summary of the `lowbirthweight` variable in the `nc` data set.
- Copy and paste the code below to add row/column totals to the table:

`addmargins(table(nc$lowbirthweight))`

- Click 'Submit Answer' and use the table to find the proportion of low birthweights.
*** =hint
Remember, to make a table use the format: `table(dataset$variable)`.

*** =pre_exercise_code
```{r}
download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")

```

*** =sample_code
```{r}
# Type table(dataset$categ_var) with specified data set and variable to obtain a table summary


# Copy and paste addmargins(table(nc$lowbirthweight)) to get margin totals



```

*** =solution
```{r}
# Type table(dataset$categ_var) with specified data set and variable to obtain a table summary

table(nc$lowbirthweight)

# Copy and paste addmargins(table(nc$lowbirthweight)) to get margin totals

addmargins(table(nc$lowbirthweight))



```

*** =sct
```{r}

test_function("table", args = "...", incorrect_msg = "Follow the format in the lesson with correct data set (`nc`) and categ_var (`lowbirthweight`).")

test_student_typed("addmargins(table(nc$lowbirthweight))", not_typed_msg = "Don't forget to copy and paste the code from the 2nd instruction!")


success_msg("Note that the 'low' birthweight column is on the left. Use the table from the output to find the proportion of low birthweights by dividing the number of low birthweights by the sum.")


```


--- type:NormalExercise lang:r xp:100 skills:1 key:5d43584cb1
## One sample z-test with prop.test()

We can see that the proportion of babies with low birthweights from this data set is around 11%.

However, a previous study estimated the low birthweight rate to be 7%. In other words, 7% of all births are under weight. 

If we want to test whether the low birthrate of our population differs from 7%, we can use the `prop.test()` function with the format:

`prop.test(x = #yes, n = #total, p = null_hyp, correct = FALSE)`

* *#yes* = number of births classified as low birthweight
* *#total* = total number of births in the sample
* *null_hyp* = the proportion from our null hypothesis


Our hypotheses are:

$H_0$: $p \= 0.07$

$H_a$: $p \neq 0.07$

where $p$ = true proportion of low birthweight births

Remember, there were 111 babies with low birthweight and 1000 total births in the sample.

*** =instructions
Use the `prop.test()` function to perform a *z*-test to determine if the population low birthweight rate differs from 0.07 (or 7%).
Remember, we are using the `nc` data set and `lowbirthweight` variable.


*** =hint

Remember to set p = sample proportion (in this case `p = 0.07`).

*** =pre_exercise_code
```{r}
download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")

```

*** =sample_code
```{r}
# use the prop.test() function with the format from the lesson to test if the the population proportion differs from 0.07.



```

*** =solution
```{r}
# use the prop.test() function with the format from the lesson to test if the the population proportion differs from 0.07.
prop.test(x = 111, n = 1000, p = 0.07, correct = FALSE)


```

*** =sct
```{r}
test_function("prop.test", args = c("x", "n", "p", "correct"))
success_msg("Great! Now examine the R output. The p-value is small so we can reject the null hypothesis and conclude the proportion of low birthweights is significantly different than 0.07.")


```

--- type:NormalExercise lang:r xp:100 skills:1 key:b187ce8175
## Two-way contingency table

The `nc` data set also contains a categorical variable called `habit` with 2 levels:

- the mother is a smoker (`smoker`)
- the mother is not a smoker (`nonsmoker`)

Suppose we want to test whether low birthweight rates vary depending on the mothers smoking habits.

Let's begin by obtaining a table of `lowbirthweight` by `habit`.

To create this contingency table, use the `table()` function with the format:

`table(dataset$var1, dataset$var2)`


*** =instructions
- Create a table of `lowbirthweight` and `habit` from the `nc` data set where:

* var1 = `lowbirthweight`
* var2 = `habit`

- Copy and paste the code below to add row/column totals to the table:

`addmargins(table(nc$lowbirthweight, nc$habit))`

- Click the 'Submit Answer' button.


*** =hint
- For the first instruction, follow the format in the lesson with correct data set (`nc`), var1 (`lowbirthweight`) and var2 (`habit`).
*** =pre_exercise_code
```{r}
download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")

```

*** =sample_code
```{r}
# Create a table of low birthweights and smoking habits using the table() function


# Copy and paste the code from the second instruction


```

*** =solution
```{r}
# Create a table of low birthweights and smoking habits using the table() function
table(nc$lowbirthweight, nc$habit)


# Copy and paste the code from the second instruction
addmargins(table(nc$lowbirthweight, nc$habit))


```

*** =sct
```{r}
test_function("table", args = "...", eval = TRUE, incorrect_msg = "Follow the format in the lesson with correct dataset (`nc`), var1 (`lowbirthweight`) and var2 (`habit`).")

test_student_typed("addmargins(table(nc$lowbirthweight, nc$habit))", not_typed_msg = "Don't forget to copy and paste the code from the 2nd instruction!")

success_msg("Good job.")

```


--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:c371bb4088
## Quick check 1

The contingency table from the last exercise is stored in an object called `table1`; type `table1` in your R script to view it.

What is the low birthweight rate among smoking mothers in this sample?

*** =instructions
- 16.4%
- 10.5%
- 14.3%
- 85.7%
*** =hint

*** =pre_exercise_code
```{r}
download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")

table1 <- addmargins(table(nc$lowbirthweight, nc$habit))
```

*** =sct
```{r}
msg_bad <- "That is not correct. Use `table1` to find the proportion of low birthweights by dividing the number of low birthweights among smoking mothers by the total number of smoking mothers."
msg_success <- "Exactly! The low birthweight rate is higher among smoking mothers (as expected) but let's see if this difference is meaningful beyond random chance!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:a51b77b946
## Two sample z-test with prop.test()

To test whether there is a significant difference between 2 population proportions, we can use the `prop.test()` function with the format:

`prop.test(c(x1, x2), c(n1, n2), correct = FALSE)` with:

* *x1* = number low birthweight babies among smoking mothers (18)
* *x2* = number low birthweight babies among nonsmoking mothers (92)
* *n1* = number of smoking mothers (126)
* *n2* = number of nonsmoking mothers (873)

In this exercise you will perform a 2 sample z-test.

Our hypotheses are:

$H\_0$: $p\_{smoker} = p\_{nonsmoker}$

$H\_a$: $p\_{smoker} \neq p\_{nonsmoker}$
****
$p\_{smoker}$ = proportion of low birthweight babies among smoking mothers

$p\_{nonsmoker}$ = proportion of low birthweight babies among nonsmoking mothers

*** =instructions
Use the `prop.test()` function to test whether the population proportion of low birthweight babies differs by mother's smoking status. 

*** =hint
Follow the format from the lesson with the correct arguements (x1 = 18, x2 = 92, n1 = 126, n2 = 873).
*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}
# Use the prop.test() function with the format from the lesson to perform a 2 sample z-test



```

*** =solution
```{r}
# Use the prop.test() function with the format from the lesson to perform a 2 sample z-test
prop.test(c(18, 92), c(126, 873), correct = FALSE)


```

*** =sct
```{r}
test_function("prop.test", args = c("x", "n", "correct"))
success_msg("Great! Now look at the R output to see the results of the test.")


```


--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:1d20719892
## Test conclusion

The 2 sample *z*-test from the last exercise are stored in an object called `test_results`; in your R console type `test_results` to view these.

At the $\alpha = 0.05$ significance level, how would we conclude this test?

Recall our hypotheses are:

$H\_0$: $p\_{smoker} = p\_{nonsmoker}$

$H\_a$: $p\_{smoker} \neq p\_{nonsmoker}$

*** =instructions
- We would reject the null hypothesis and conclude that there is a significant difference in the proportion of smokers and nonsmokers with low birthweight births.
- We would fail to reject the null hypothesis and conclude that there is no evidence of a difference in the proportion of smokers and nonsmokers with low birthweight babies.

*** =hint

Remember, if zero is in the confidence interval it is plausible that the true difference in proportions is zero.

*** =pre_exercise_code
```{r}
test_results <- prop.test(c(18, 92), c(126, 873), correct = FALSE)
```

*** =sct
```{r}
msg_bad <- "That is not correct. The p-value is greater than 0.05 AND the confidence interval contains zero so we would fail to reject the null hypothesis."
msg_success <- "Exactly!"
test_mc(correct = 2, feedback_msgs = c(msg_bad, msg_success))

```


--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:c971d0fe82
## Quick check 2

Suppose I get a random sample of 200 Cal Poly students, and I find that there are 40 athletes and 160 non-athletes. Among the 40 athletes, 15 were on the Dean's List. Among the 160 non-athletes, 33 were on the Dean's List. I want to use this data to test if the population proportion of all Cal Poly athletes that are on the Dean's List differs from the population proportion of all Cal Poly non-athletes that are on the Dean's List.

Which of the following commands would correctly execute this test?

*** =instructions
- prop.test(c(15,40),c(33,160),correct=F)
- prop.test(c(15,160),c(33,40),correct=F)
- prop.test(c(15,33),c(40,160),correct=F)
- prop.test(c(15,200),c(33,200),correct=F)

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```

--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:770e869325
## Quick Check 3

The two sample z-test is appropriate when we want to:
*** =instructions
- examine two paired (dependent) categorical variables
- compare more than two population proportions
- compare two population proportions
- test a single population proportion
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. With a 2 sample z-test we are dealing with 2 independent proportions."
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))


```





--- type:NormalExercise lang:r xp:100 skills:1 key:cfdac30423
## Exploring two categorical variables 

What if we want to compare proportions of *more than* 2 groups? To do this we can do a chi-squared test of association.

To perform a chi-squared test, we'll be using the same data set from Lab 5 called `HELPrct` which contains data on adult inpatients who were recruited from a detoxification unit. 

- there are 27 variables in this data set but we will only use `homeless` and `substance`
- there are 453 subjects 
- `HELPrct` is already in your workspace

`homeless` is a categorical variable with 2 levels:

- *homeless* (the patient was homeless before their detoxification stay)
- *housed* (the patient was housed before their detoxification stay)

`substance` is a categorical variable indicating the patients' primary substance of abuse with 3 levels:

- *alcohol*
- *cocaine*
- *heroin*

We want to test whether the homeless rate varies by primary substance type. Now we have **3** groups of interest so we need to perform this test with the chi-squared test of association.

Let's begin by exploring the `homeless` and `substance` variables in the `HELPrct` data set.


*** =instructions
- Use the `table()` function to create a table of `homeless` and `substance` where:

* var1 = `homeless`
* var2 = `substance`

- Copy and paste the code below to add row/column totals to the table:

`addmargins(table(HELPrct$homeless, HELPrct$substance))`

- Click the 'Submit Answer' button and look at the R output.
*** =hint
To create this contingency table, use the `table()` function with the format:

`table(dataset$var1, dataset$var2)`

*** =pre_exercise_code
```{r}
library(mosaicData)
```

*** =sample_code
```{r}
# Create a table of homeless and substance using the table() function


# Copy and paste the code from the second instruction


```

*** =solution
```{r}
# Create a table of homeless and substance using the table() function
table(HELPrct$homeless, HELPrct$substance)

# Copy and paste the code from the second instruction
addmargins(table(HELPrct$homeless, HELPrct$substance))


```

*** =sct
```{r}
test_function("table", args = "...", eval = TRUE, incorrect_msg = "Follow the format in the lesson with correct dataset (`HELPrct`), var1 (`homeless`) and var2 (`substance`).")

test_student_typed("addmargins(table(HELPrct$homeless, HELPrct$substance))", not_typed_msg = "Don't forget to copy and paste the code from the 2nd instruction!")

success_msg("Good work.")


```

--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:87d9297923
## Sample proportions

The contingency table from the last exercise is stored in an object called `my_table`; type `my_table` in your R script to view it.

Which group has the highest homeless rate?

*Note: you can use R to calculate the proportions.*
*** =instructions
- heroin users
- cocaine users
- alcohol users

*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
my_table <- addmargins(table(HELPrct$homeless, HELPrct$substance))

```

*** =sct
```{r}
msg_bad <- "That is not correct. Find the homeless rates by dividing the number of homeless people in each substance group by the total number of people in that substance group."
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:4a650270a2
## Chi-squared test

To perform a chi-squared test of association we can use the `chisq.test()` function with the format:

`chisq.test(dataset$var1, dataset$var2, correct = FALSE)`

For this exercise:

* var1 = `homeless`
* var2 = `substance`


*** =instructions
Use the `chisq.test()` function to test whether the homeless rate differs depending on the patients' primary substance (from the `HELPrct` data set).
*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
```

*** =sample_code
```{r}
# Perform a chi-squared test on homeless rate by substance 



```

*** =solution
```{r}
# Perform a chi-squared test on homeless rate by substance 
chisq.test(HELPrct$homeless, HELPrct$substance, correct = F)



```

*** =sct
```{r}
test_function("chisq.test", args = c("x", "y", "correct"), args_not_specified_msg = "Make sure you follow the format from the lesson and don't forget to add 'correct = FALSE'. Remember, the data set we are using is called `HELPrct`.")

success_msg("Great! Now look at the R output to see the results of the test.")

```





--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:ca5a31f85a
## Chi-squared test conclusion

Our hypotheses for the last exercise:

$H_0$: Homelessness is not associated with primary substance.

$H_a$: Homelessness is associated with primary substance.


The chi-squared test from the last exercise re stored in an object called `chisq_results`; in your R console type `chisq_results` to view these.

At $\alpha$ = 0.05, we ______ the null hypothesis and conclude that we ____ have evidence of an association between homelessness and primary substance.


*** =instructions
- reject / do
- reject / do not
- fail to reject / do
- fail to reject / do not

*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
chisq_results <- chisq.test(HELPrct$homeless, HELPrct$substance, correct = F)
```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))

```



--- type:NormalExercise lang:r xp:100 skills:1 key:140356f17c
## Assumptions of the chi-squared test

An assumption of the chi-squared test is that all expected counts are at least 5.

To check this, you will save the chi-squared test results as an object and then use `$expected` to view the expected cell counts.

*** =instructions
- Copy and paste the code below to save the test results in the object called `results`.

`results <- chisq.test(HELPrct$homeless, HELPrct$substance, correct=F)`

- View the expected counts by typing `results$expected`.

- Click the 'Submit Answer' button and make sure all of the cell counts are over 5.
*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
```

*** =sample_code
```{r}
# Copy and paste the code from the first instruction


# Access the expected cell counts by typing 'results$expected'



```

*** =solution
```{r}
# Copy and paste the code from the first instruction
results <-chisq.test(HELPrct$homeless, HELPrct$substance,correct=F)

# Access the expected cell counts by typing 'results$expected'
results$expected


```

*** =sct
```{r}
test_object("results")
success_msg("Nice job. Since all of the expected counts are at least 5, this condition is satisfied for valid inference.")


```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:5cefd02043
## LAST EXERCISE OF LAB 7

**This is the last exercice of Lab 7.**

When we do the one sample z-test or the two sample z-test we are interested in doing statistical inference about

*** =instructions
- only means
- only proportions 
- both means and proportions
- neither means nor proportions

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly! You have completed Lab 7 practice course."
test_mc(correct = 2, feedback_msgs = c(msg_bad, msg_success, msg_bad, msg_bad))

```

