---
title       : Lab 8
description : Correlation and Simple Linear Regression



--- type:NormalExercise lang:r xp:100 skills:1 key:283c94164f
## Gun violence data

In this pre-lab, we'll be using a data set called `gun_violence_us` which contains gun ownership and mortality statistics in the United States.

- there are 3 variables in this data set: `state`, `ownership_rate`, `mortality_rate`
- `gun_violence_us` is already in your workspace

Let's begin by exploring the data set.
*** =instructions
- Use the `head()` function on `gun_violence_us` to view the first six observations.
- Use the `summary()` function on `gun_violence_us` to obtain summary statistics for each variable.
*** =hint
Use the format `head(dataset)` for the first instruction and `summary(dataset)` for the second instruction.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/gun_violence_us.R")

```

*** =sample_code
```{r}
# Use the `head()` function on `gun_violence_us` to view the first six observations.


# Use the `summary()` function on `gun_violence_us` to obtain summary statistics for each variable.


```

*** =solution
```{r}
# Use the `head()` function on `gun_violence_us` to view the first six observations.
head(gun_violence_us)

# Use the `summary()` function on `gun_violence_us` to obtain summary statistics for each variable.
summary(gun_violence_us)

```

*** =sct
```{r}
test_function("head", args = "x", incorrect_msg = "Make sure to use 'gun_violence_us' instead of 'dataset' inside the head() function.")

test_function("summary", args = "object", incorrect_msg = "Follow the format 'summary(dataset)' with the specified data set to obtain a summary of the entire data set.")

```

--- type:NormalExercise lang:r xp:100 skills:1 key:b81533a6c6
## Correlation

Suppose we want to test whether there is a relationship between `ownership_rate` and `mortality_rate` in the `gun_violence_us` data set.

* `ownership_rate`: percentage of adults in each state who own a gun

* `mortality_rate`: number of deaths per 100,000 people 

Let's start by visualizing the relationship between these variables with the `plot()` function.

Remember, to create a scatterplot use the format:

`plot(dataset$x.var, dataset$y.var)`
*** =instructions
Use the `plot()` function to visualize the relationship between `ownership_rate` and `mortality_rate` where:

* x.var = `ownership_rate`
* y.var = `mortality_rate`
*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/gun_violence_us.R")
```

*** =sample_code
```{r}
# Use the `plot()` function to visualize the relationship between `ownership_rate` and `mortality_rate`


```

*** =solution
```{r}
# Use the `plot()` function to visualize the relationship between `ownership_rate` and `mortality_rate`
plot(gun_violence_us$ownership_rate, gun_violence_us$mortality_rate)


```

*** =sct
```{r}
test_function("plot", args = c("x", "y"), incorrect_msg = "Remember to follow the format in the lesson with the specified data set (gun_violence_us) and variables.")

success_msg("Nice! Now look at the plot. We can see that as gun ownership rates increase, mortality rates also increase.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:23b4d73093
## Measuring correlation with the cor() function

We can use the `cor()` function to estimate the correlation between two quantitative variables with the format:

`cor(dataset$var1, dataset$var2)`

*Note: the order of var1 and var2 does not matter for the `cor()` function.*

*** =instructions
Use the `cor()` function to estimate the correlation coefficient between `ownership_rate` and `mortality_rate` from the `gun_violence_us` data set.
*** =hint
Follow the format `cor()` format in the lesson with the `gun_violence_us` data set and `ownership_rate`/`mortality_rate` in any order. 
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/gun_violence_us.R")
```

*** =sample_code
```{r}
# Estimate the correlation between ownership_rate and mortality_rate


```

*** =solution
```{r}
# Estimate the correlation between ownership_rate and mortality_rate
cor(gun_violence_us$ownership_rate, gun_violence_us$mortality_rate)


```

*** =sct
```{r}
test_function("cor", args = c("x", "y"), eval = NA)
```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:e07efd5860
## Quick check 1

Using the command `cor(x, y)` will produce different results than the command `cor(y, x)`.
*** =instructions
- TRUE
- FALSE
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_bad, msg_success))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:76ae5ee967
## Simple linear regression

We can learn more about the relationship between two variables by calculating a linear model with the `lm()` function in the format:

`model_name <- lm(dataset$y.var ~ dataset$x.var)`

`model_name`

* y.var = the response variable
* x.var = the explanatory variable


Note: Remember that after we create an object we need to type the object name to view it *(I have already typed this for you in the R Script)*.

*** =instructions
* Use the `lm()` function to calculate a linear model from the `gun_violence_us` data set and store these regression results in an object called `gun_model` where:

    - x.var = `ownership_rate`
    - y.var = `mortality_rate`

*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/gun_violence_us.R")
```

*** =sample_code
```{r}
# Use the lm() function to calculate a linear model of ownership_rate and mortality_rate
gun_model <- 

gun_model


```

*** =solution
```{r}
# Use the lm() function to calculate a linear model of ownership_rate and mortality_rate
gun_model <- lm(gun_violence_us$mortality_rate ~ gun_violence_us$ownership_rate)

gun_model

```

*** =sct
```{r}
test_object("gun_model", incorrect_msg = "Use the lm() function on the right side of the tilde with correct data set (gun_violence_us), y.var = mortality_rate and x.var = ownership_rate.")

success_msg("Good work. Now look at the R output and find the estimate for the intercept and slope.")
```





--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:649d942063
## Quick check 2

In the last exercise, the estimate for the intercept was 4.573 and the estimate for the slope was 20.492.

What is our linear regression model?

*** =instructions
- $\hat{y} = 4.573 + 20.492x$
- $\hat{y} = 4.573 + 20.492$
- $\hat{y} = 20.492 + 4.573x$
- $\hat{y} = 20.492 + 20.492x$

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:3605a42869
## Testing significance of correlation

Suppose we want to test whether there is a relationship between `ownership_rate` and `mortality_rate` in the `gun_violence_us` data set.

Our hypotheses:

$H_0: \beta = 0$

$H_a: \beta \neq 0$

where $\beta$ = the slope of the regression

*Note: `gun_model` is already in your workspace.*
*** =instructions
- Summarize `gun_model` using the `summary()` function.
- Click the 'Submit Answer' button and look at the p-value of the test.
*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/gun_violence_us.R")
gun_model <- lm(gun_violence_us$mortality_rate ~ gun_violence_us$ownership_rate)

```

*** =sample_code
```{r}
# Use the summary function on gun_model


```

*** =solution
```{r}
# Use the summary function on gun_model
summary(gun_model)


```

*** =sct
```{r}
test_function("summary", args = "object")
success_msg("Good job. The p-value for this test is almost zero so we can reject the null hypothesis and conclude that the slope is significantly greater than zero.")
```



--- type:NormalExercise lang:r xp:100 skills:1 key:cded916a25
## Add lines to a plot with abline()

We can add the estimated regression equation to the scatter plot with the `abline()` function with the format:

`abline(model_name)`

*** =instructions
- The first line of code creates a scatter plot of `ownership_rate` and `mortality_rate` from the `gun_violence_us` data set. **Do not alter this code.**

- Use the `abline()` function to add the `gun_model` regression line to the plot.
*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/gun_violence_us.R")
gun_model <- lm(gun_violence_us$mortality_rate ~ gun_violence_us$ownership_rate)

```

*** =sample_code
```{r}
# Code to create a scatter plot of ownership_rate and mortality_rate. Do not change
plot(gun_violence_us$ownership_rate, gun_violence_us$mortality_rate)


# Add the gun_model regression line to the scatter plot


```

*** =solution
```{r}
# Code to create a scatter plot of ownership_rate and mortality_rate. Do not change
plot(gun_violence_us$ownership_rate, gun_violence_us$mortality_rate)


# Add the gun_model regression line to the scatter plot
abline(gun_model)


```

*** =sct
```{r}
test_student_typed("plot(gun_violence_us$ownership_rate, gun_violence_us$mortality_rate)", not_typed_msg = "Do not remove or alter the line of code that plots ownership_rate and mortality_rate. Instead, do the second instruction.")

test_function("abline", args = "a", incorrect_msg = "Make sure to follow the format in the lesson with gun_model instead of model_name.")

```

--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:ecd40fc973
## Quick check 3

Correlation can be used to estimate the strength of association between:

*** =instructions
- one quantitative variable and $\mu_{0}$
- two quantitative variables
- two categorical variables
- one quantitative and one categorical variable

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. What kind of variables were 'ownership_rate' and 'mortality_rate'?"
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_bad, msg_success, msg_bad, msg_bad))

```
