---
title       : Lab 2
description : Using R to Explore Data Sets

--- type:NormalExercise lang:r xp:100 skills:1 key:f8b383874f
## Getting started with data

For this lab, we will use a dataset called `arbuthnot` which contains data on male and female births in London from 1629-1710.

- `arbuthnot` is already in your workspace so if you type `arbuthnot` in your console, the data set will print.

Sometimes the data can be very long so instead of viewing the entire thing, we can type:

- `head(dataset)` to get the first 6 entries or
- `tail(dataset)` to get the last 6 entries. 

If we want to know the variable (column) names in the data set, we can type `names(dataset)` into the console.

If we want to know the dimensions of a dataset (ie. how many variables and subjects) use the `dim(dataset)` function.

Note: `dataset` is just a placeholder for the name of the data set you want to explore. In this exercise, you need to substitute `arbuthnot` for `dataset`.

*** =instructions
- Use the `head()` function to view the first 6 entries of the `arbuthnot` data set.
- Use the `tail()` function to view the last 6 entries.
- Find the variable names by using the `names()` function.
- Use `dim()` to find the dimensions of `arbuthnot`.
- Click the 'Submit Answer' button and take a look at the R output in the console.


*** =hint
The data set is named `arbuthnot` so substitute `arbuthnot` for `dataset`.

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
#Use appropriate function to find first 6 entries of arbuthnot

#Use appropriate function to find last 6 entries of arbuthnot

#Use appropriate function to find the variable names of arbuthnot

#Use appropriate function to find the dimension of arbuthnot


```

*** =solution
```{r}
#Use appropriate function to find first 6 entries of arbuthnot
head(arbuthnot)

#Use appropriate function to find last 6 entries of arbuthnot
tail(arbuthnot)

#Use appropriate function to find the variable names of arbuthnot
names(arbuthnot)

#Use appropriate function to find the dimension of arbuthnot
dim(arbuthnot)

```

*** =sct
```{r}

test_function("head", args = "x", incorrect_msg = "Make sure to use 'arbuthnot' instead of 'dataset' inside the head() function.")


test_function("tail", args = "x", incorrect_msg = "Make sure to use 'arbuthnot' instead of 'dataset' inside the tail() function.")


test_function("names", args = "x", incorrect_msg = "Make sure to use 'arbuthnot' instead of 'dataset' inside the names() function.")


test_function("dim", args = "x", incorrect_msg = "Make sure to use 'arbuthnot' instead of 'dataset' inside the dim() function.")

success_msg("Good work! Now look at the R output, notice how each horizontal row is a different year. We can see from the dim() command that there are 82 years with 3 variables.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:795745ad44
## Review accessing variables

Remember, to access specific variables within a data set use the format:

`dataset$variable`.

You might have seen in the previous exercise that `arbuthnot` has a *number of male births* variable named `boys`. If we want to look at this variable by itself, we can type `arbuthnot$boys` into the R console; this will return the number of male births each year. 


*** =instructions
- The arbuthnot data set also has a *number of female births* variable named `girls`. Use `$` to access the `girls` variable from the `arbuthnot` data set.
*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Access girls from arbuthnot like we did for the boys variable


```

*** =solution
```{r}
# Access girls from arbuthnot like we did for the boys variable
arbuthnot$girls
```

*** =sct
```{r}
test_student_typed("arbuthnot$girls", not_typed_msg = "Make sure you follow the format: dataset$variable with the correct data set and variable!")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:f759511088
## Using the summary() function

In this exercise, you will learn how to use the R built-in function `summary()` to obtain statistics of your data set. 

`summary(dataset)`: gives a few summary statistics of all variables in dataset


`summary(dataset$variable)`: gives a few summary statistics of a specific variable in dataset

The dataset we've been working with `arbuthnot` is already in your workspace.

*** =instructions
- Use the `summary()` function on the entire `arbuthnot` dataset.

*** =hint
substitute `arbuthnot` for dataset

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Use summary() on arbuthnot to get summary statistics of the entire arbuthnot dataset
```

*** =solution
```{r}
# Use summary() on arbuthnot to get summary statistics of the entire arbuthnot dataset
summary(arbuthnot)

```

*** =sct
```{r}
test_function("summary", args = "object", incorrect_msg = "Follow the format 'summary(dataset)' to obtain a summary of the entire data set.")
```

--- type:NormalExercise lang:r xp:100 skills:1 key:15bd840872
## Working with Variables **this needs work

We can also use arithetic operators on variables. 

For example, if I want to add two variables together, i would type `dataset$var1 + dataset$var2`. 

If I want to save this sum as a new variable, I would assign it to a name `dataset$newvar <- dataset$var1 + dataset$var2`.

In this exercise, we will be creating a new variable called `propBoys` which measures the proportion of male births each year.

*propBoys <- #boys/#total*

To do this, we will use the formula:

`dataset$newvar <- dataset$var1 / (dataset$var1 + dataset$var2)`


*** =instructions
- Use the formula above to create the `propBoys` variable where:
    - dataset = `arbuthnot`
    - newvar = `propBoys`
    - var1 = `boys`
    - var2 = `girls`
- Use the `names()` function on `arbuthnot` to make sure that `propBoys` was added to the variables.
*** =hint
Follow the formula in the instructions with specified variables and data set (start typing on the right side of `<-`.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Create propBoys variable for arbuthnot
arbuthnot$propBoys <- 

# Use names() to check if propBoys was added


```

*** =solution
```{r}
# Create propBoys variable for arbuthnot
arbuthnot$propBoys <- arbuthnot$boys / (arbuthnot$boys + arbuthnot$girls)

# Use names() on arbuthnot to check if propBoys was added
names(arbuthnot)
```
*** =sct
```{r}

success_msg("Great. Now look at the output and make sure that `propBoys` was added to the variables.")
```



--- type:NormalExercise lang:r xp:100 skills:1 key:59b987e2d7
## Using the sum() function

Another useful built-in function that R has is the `sum()` function. 

The `sum()` function adds what you put into the parenthesis.

For example, to add a column in a data set use the format:

`sum(dataset$variable)`

*** =instructions
Use the `sum()` function to find the total number of male births `boys` in `arbuthnot`.
*** =hint
- To sum over a variable, use the format: `sum(dataset$variable)`.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Find total number of boys using the sum() function



```

*** =solution
```{r}
# Find total number of boys using the sum() function
sum(arbuthnot$boys)


```

*** =sct
```{r}
test_function("sum", args = "...", incorrect_msg = "Make sure you use the format: sum(dataset$variable) with the correct data set and variable!")


```



--- type:NormalExercise lang:r xp:100 skills:1 key:938ccd0b92
## Creating Logical Data

Recall in Lab 1 when we reviewed the different types of data.

*Logical* data only takes on two values (TRUE or FALSE) where `TRUE` has the value of 1 and `FALSE` has the value of 0.

We can create *logical* data with comparisons like greater than `>`, less than `<`, and inequallity `!=`

A logical expression using the *greater than* symbol should follow the format:

`dataset$var1 > dataset$var2`

This command will return a list of `TRUE` and `FALSE` values.

We can save these values in an object `new_var` by typing:

`dataset$new_var <- dataset$var1 > dataset$var2`
*** =instructions
- Find which years in `arbuthnot` had greater male births `boys` than female births `girls`, save these values by assigning them to a new variable named `my_var`.
- Use the sum() function to evaluate the total number of `TRUE` results in `my_var`.
*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

*** =sample_code
```{r}
# Use the code at the end of the lesson with correct data set and variables


# Use the sum() function on my_var to find how many years there were more male births than female births


```

*** =solution
```{r}
# Use the code at the end of the lesson with correct data set and variables
arbuthnot$my_var <- arbuthnot$boys > arbuthnot$girls


# Use the sum() function on my_var to find how many years there were more male births than female births
sum(arbuthnot$my_var)

```

*** =sct
```{r}

test_student_typed("arbuthnot$my_var <- arbuthnot$boys > arbuthnot$girls", not_typed_msg = "That is not correct. Follow the format in the lesson with 'arbuthnot' instead of 'dataset', 'my_var' instead of 'new_var', 'boys' instead of 'var1', and 'girls' instead of 'var2'")

test_function("sum", args = "...", incorrect_msg = "Remember, the format for finding the sum of a variable is 'sum(dataset$variable)'. In this example, the dataset is 'arbuthnot' and the variable is 'my_var'.")

```



--- type:NormalExercise lang:r xp:100 skills:1 key:6bc90b9c68
## Graphing Data: the Basics

Numeric summaries like the mean and standard deviation are helpful when trying to understand a dataset but a visual display can provide more insight. 

To create a basic plot in R, use the `plot()` function in the format:

`plot(x = dataset$var1 , y = dataset$var2)`

For this exercise, we are going to visualize the relationship between two variables in `arbuthnot`:

x = year (`year`) 

y = proportion of male births (`propBoys`)



*** =instructions
- Use the `plot()` function to create a scatterplot of `propBoys` by `year` from the `arbuthnot` data set.
- After clicking 'Submit Answer', look at the plot in the output. Is there any pattern?


*** =hint

*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
arbuthnot$propBoys <- arbuthnot$boys / (arbuthnot$boys + arbuthnot$girls)

```

*** =sample_code
```{r}
# Use the plot() function with x = arbuthnot$year and y = arbuthnot$propBoys


```

*** =solution
```{r}
# Use the plot() function with x = arbuthnot$year and y = arbuthnot$propBoys

plot(x = arbuthnot$year, y = arbuthnot$propBoys)


```

*** =sct
```{r}
test_function("plot", args = c("x", "y"), incorrect_msg = "Make sure you follow the formula in the lesson with the correct data set and variables!")
success_msg("Good work!")



```


--- type:NormalExercise lang:r xp:100 skills:1 key:d0401021d4
## Adding axis labels to graphs 

In this exercise, you will learn how to clean up a graph by adding axis labels and a title.

Some arguements to specify in the `plot()` function:

- `xlab = "x"` : labels the x-axis "x"
- `ylab = "y"` : labels the y-axis "y"
- `main = "title"` : gives a title to the entire plot "title"

To add labels to a plot use the format:

`plot(x = dataset$var1 , y = dataset$var2, xlab = "x", ylab = "y", main = "title")`

Right now the plot of `arbuthnot$year` vs. `arbuthnot$propBoys` has ugly axis names. In this exercise, we will change the axis names so the plot looks less messy.

*** =instructions
- Create the plot from the last exercise `plot(x = arbuthnot$year, y = arbuthnot$propBoys)` but this time add axis labels. 

    - The new x-axis label should be `"Year"`, y-axis label should be `"Proportion of Male Births"` and the title should be `"Arbuthnot: Boys by Year"`.
- Click the 'Submit Answer' button and take a look at the new plot. So much Better!

*** =hint
Make sure to wrap the labels in "" because they are words.
*** =pre_exercise_code
```{r}
source("http://www.openintro.org/stat/data/arbuthnot.R")
arbuthnot$propBoys <- arbuthnot$boys / (arbuthnot$boys + arbuthnot$girls)

plot(x = arbuthnot$year, y = arbuthnot$propBoys)
```

*** =sample_code
```{r}
# Use plot(x = dataset$var1 , y = dataset$var2, xlab = "x", ylab = "y", main = "title") with specified labels



```

*** =solution
```{r}
# Use plot(x = dataset$var1 , y = dataset$var2, xlab = "x", ylab = "y", main = "title") with specified labels

plot(x = arbuthnot$year , y = arbuthnot$propBoys, xlab = "Year", ylab = "Proportion of Male Births", main = "Arbuthnot: Boys by Year")


```

*** =sct
```{r}
test_function("plot", args = c("x", "y", "xlab", "ylab", "main"), incorrect_msg = "Use the format in the lesson with the specified lables in the instructions. Make sure you wrap the labels in quotation marks because they are words.")
success_msg("Look how much better the graph looks now that we have labels! You have completed the Lab 2 practice course.")
```
