---
title       : Lab 1
description : Introduction to R and RStudio




--- type:NormalExercise lang:r xp:100 skills:1 key:8902293ef6
## How to use DataCamp

The upper right portion of your screen is the script editor: this is where you will be typing your R code to answer the questions. When you press the 'Submit Answer' button, every line of code is interpreted and executed by R and you will get a message that says whether or not your code was correct. 

Below the script is the R Console: you can use this to execute R commands and test your R code for correctness before you submit it. 

The # symbol lets you comment on your code without R processing it.

*** =instructions 
This exercise is only meant to help you get set up with DataCamp so there is no coding task.
Just click 'Submit Answer' to continue to the next exercise.
*** =hint

*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}
# This is were you will be writing your code



```

*** =solution
```{r}
# This is were you will be writing your code



```

*** =sct
```{r}

```
--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:77ff5d2cfd
## Getting started with R


The best way to learn R programming is to actually practice typing yourself. 
So, lets start coding!

Are you excited to dive into the exhilarating world of R Programming?!

*** =instructions
- Yes
- No

*** =hint
You should be very excited!
*** =pre_exercise_code
```{r}

```


*** =sct
```{r}
msg_bad <- "Well, hopefully it will grow on you!"
msg_success <- "Good answer!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad))
```

--- type:NormalExercise lang:r xp:100 skills:1 key:dbdb4a1084
## Using R as a calculator


In this exercise, we'll see how R can be used to perform simple arithmetic tasks.

Some arithmetic operators you can use in R code include:

Addition: `+`

Subtraction: `-` 

Division: `/` 

Multiplication: `*` 

Exponentiation: `^`

For example, if I want to cube '4', I would type `4^3` into the console.

Note: When multiplying two things together use an asterisk $\ast$. For example if we want to multiply 4 and 8, type `4*8`; typing `4(8)` will result in an error.

*** =instructions
- In the script, type `4^3` to calculate four to the power of three.
- Click the 'Submit Answer' Button and take a look at the R output in the console (bottom right panel).
*** =hint
- Type your code in the upper right box.


*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}
# Example of Addition
5 + 10

# Example of division and subtraction
(10 - 5) / 2

# Use R as a calculator to find four to the power of three.




```


*** =solution
```{r}
# Example of Addition
5 + 10

# Example of division and subtraction
(10 - 5) / 2

# Use R as a calculator to find four to the power of three
4^3



```

*** =sct
```{r}
test_output_contains("15", incorrect_msg = "Do not remove the line of R code that calculates the sum of 5 and 10. Instead, just do the last question.")


test_output_contains("2.5", incorrect_msg = "Do not remove the line of R code that calculates (10 - 5) / 2. Instead, just do the last question.")

test_output_contains("64", incorrect_msg = "Make sure to add a line of R code, that calculates 4 to the power of 3. Do not start the line with a `#`, otherwise, your R code will not be executed!")

success_msg("Awesome! See how the console shows the result of the R code you submitted? Now that you're familiar with the interface, let's get down to R business!")
```



--- type:NormalExercise lang:r xp:100 skills:1 key:b19e862406
## Using the sqrt() function

Like any other calculator, R has the ability to take the square root of non-negative numbers.

We will be calculating the square root a lot in this class so knowing this R function is really handy.

If I want to take the square root of 9, I would type `sqrt(9)` and it would return 3.

*** =instructions
- In the script, use the `sqrt()` function to find the square root of 9659.
- In the script, use the `sqrt()` function and arithmetic operators (from the last exercise) to evaluate the expression:
$\frac{0.6 - 0.5}{\sqrt{0.25 / 1000}}$

*Note: R uses PEMDAS order of operations like you would use in any math class (Parentheses - exponents/square roots - multiplication - ...).*

*** =hint
R uses PEMDAS order of operations like you would use in any math class.

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Find the square root of 9659 using sqrt()

# Evaluate the expression in the plots panel 


```

*** =solution
```{r}
# Find the square root of 9659 using sqrt()
sqrt(9659)

# Evaluate the expression in the instructions 
(0.6-0.5) / sqrt(0.25/1000)

```

*** =sct
```{r}
test_function("sqrt", args = "x", incorrect_msg = "Type `sqrt(9659)` below the first comment in the R script panel.")

test_output_contains("(0.6-0.5) / sqrt(0.25/1000)", incorrect_msg = "That is incorrect. Type `(0.6-0.5) / sqrt(0.25/1000)` below the second comment to evaluate the expression.")

success_msg("Good work!")
```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:d6699314d9
## Basic data types in R

Some data types we will be working with in this course are:

Numeric: numbers with decimals like `3.1` or `22.0`

Integers: natural numbers like `-9` and `4`, integers are also numeric data.

Logical: boolean values (`TRUE` or `FALSE`)

Character: texts and words ("this is character data")


Note: all character data should be wrapped in "".

If I typed `70.0 + 28.9` into the console, the value that R would return would be what type of data?

*** =instructions
- numeric
- character
- logical
- integer

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct, `70.0 + 28.9` does not result in a whole number so the data type must be numeric."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))
```

--- type:NormalExercise lang:r xp:100 skills:1 key:8c5519a279
## Object assignment

R is what we call an *object oriented language*. In R, we can use objects to store values, data frames, or even results from statistical analysis.

A simple example of object assignment is `x <- 5` which assigns a value 5 to an object named 'x'. We can print the contents of x by typing `x` into the R console.


*** =instructions
- Assign the value 100 to a variable named `my_object`.
- Type `my_object` into your R script to print the contents.
- Click the 'Submit Answer' button and look through the R output.

*** =hint
Type '100' on the right side of the assignment command `<-` to assign 100 to `my_object`.
*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Assign the value 100 to an object named my_object

my_object <- 


# Print the value of my_object


```

*** =solution
```{r}
# Assign the value 100 to an object named my_object

my_object <- 100


# Print the value of my_object
my_object

```

*** =sct
```{r}
test_object("my_object", incorrect_msg = "Make sure you type '100' on the right side of the assignment command `<-` to assign 100 to `my_object`.")

test_output_contains("my_object", incorrect_msg = "Remember to type `my_object` to print out the contents of the object.") 

success_msg("Good! Notice that the value 100 appears when R prints `my_object`.")
```


--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:8404e9b4ff
## Quick check 1

Which of the following commands will correctly assign the numbers 1,2,3,4 to an object named `my_vector`?

*** =instructions
- my_vector = c(1, 2, 3, 4)
- my_vector <- c(1, 2, 3, 4)
- my_vector -> c(1, 2, 3, 4)
- c(1, 2, 3, 4) <- my_vector
- c(1, 2, 3, 4) -> my_vector
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. Review the last exercise."
msg_success <- "Exactly!"
test_mc(correct = 2, feedback_msgs = c(msg_bad, msg_success, msg_bad, msg_bad, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:21fd16b05f
## Viewing data
R has a built-in dataset called `women` which contains 2 variables (`height` and `weight`) of 15 women. 

This data set is already in your workspace so if you type  `women` into your R console, the data set will come up in the output.


In this exercise you will be learning how to access specific variables within a data set by using the format:

`dataset$variable`.

For example, if you want to view the `height` variable by itself, type `women$height` into the R console.

*** =instructions
- Type `women` in the R script to print the entire data set.
- Use the `$` notation presented in the lesson to only view the `weight` variable in the `women` data set.
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Print the entire data set by typing women


# Access weight from women like we did for the height variable


```

*** =solution
```{r}
# Print the entire data set by typing women
women

# Access weight from women like we did for the height variable
women$weight
```

*** =sct
```{r}
test_output_contains("women", incorrect_msg = "Type `women` in your R script (under the first comment) to print the entire data set.")

test_output_contains("women$weight", incorrect_msg = "Make sure you follow the format: dataset$variable with the correct data set (`women`) and variable (`weight`)!")

```



--- type:NormalExercise lang:r xp:100 skills:1 key:1ae116e9d3
## Using the summary() function

In this exercise, you will learn how to use the R built-in function `summary()` to obtain statistics of your data set. 

`summary(dataset)`: gives a few summary statistics of all variables in the data set

The dataset we've been working with `women` is already in your workspace.

*** =instructions
- Use the `summary()` function to obtain summary statistics of the `women` data set.

*** =hint
substitute `women` for dataset

*** =pre_exercise_code
```{r}
```

*** =sample_code
```{r}
# Use summary() on women to get summary statistics of the entire dataset



```

*** =solution
```{r}
# Use summary() on women to get summary statistics of the entire dataset
summary(women)


```

*** =sct
```{r}
test_function("summary", args = "object", incorrect_msg = "Follow the format `summary(dataset)` with `women` instead of dataset to obtain a summary of the entire data set.")
```
--- type:NormalExercise lang:r xp:100 skills:1 key:766b6a5af5
## Graphing data in R

In this exercise, you will practice using simple graphical functions in R to visualize data.

In it's most basic form, the `plot()` function will generate a scatterplot of the x and y values that have been passed through it. 

The format to make a basic scatterplot is:

`plot(x = dataset$x.var , y = dataset$y.var)`

For this exercise, we are going to visualize the relationship between the two variables in `women`.


*** =instructions
- Follow the `plot()` function format from the lesson to create a scatterplot of `height` by `weight` from the `women` data set.
    * `height` should be on the x-axis
    * `weight` should be on the y-axis

- After clicking 'Submit Answer', look at the plot in the output. Is there any pattern?
*** =hint
Follow the format in the lesson with `women` instead of `dataset`, `height` instead of `var1` and `weight` instead of `var2`.
*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Use plot() to create scatterplot of height by weight


```

*** =solution
```{r}
# Use plot() to create scatterplot of height by weight

plot(x = women$height, y = women$weight)
```

*** =sct
```{r}
test_function("plot", args = c("x", "y"), incorrect_msg = "In your R script, follow the format: `plot(x = dataset$var1 , y = dataset$var2`) with the specified arguements. In this case, use women instead of dataset, height instead of var1 and weight instead of var2.")
success_msg("Good work! From the plot we can see that as height increases, weight decreases.")

```

--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:9707072219
## Quick check 2

What can R be used for?

*** =instructions
- Graphing data
- Analyzing data
- Calculations
- All of the above

*** =hint
R software can be used for all of these things.

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}

msg_bad <- "That is not correct"
msg_success <- "Exactly! R can do all of these things."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))
```



--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:dff314deba
## Quick check 3
**This is the last exercise of Lab 1.**

What is the correct code to access a variable in a data set?

*** =instructions
- dataset#variable.name
- dataset&variable.name
- dataset=variable.name
- dataset$variable.name
- dataset%variable.name
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. We use a dollar sign `$` to access variables."
msg_success <- "Exactly! You finished Lab 1 practice course."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success, msg_bad))


```
