---
title       : Lab 4b
description : Population and Sample Means


--- type:NormalExercise lang:r xp:100 skills:1 key:57cc190216
## Review random sampling

In this lab we'll be using another built-in R dataset called `ChickWeight` which contains data from an experiment on the effect of diet on chicks. 

- There are 4 variables in the data set: 
    - `weight`: the chicks body weight in grams
    - `Time`: when they were weighed
    - `Chick`: unique identifier for the chick
    - `Diet`: the treatment diet assigned to the chicks with varying amounts of protein and fat; either Diet 1, 2, 3, or 4
- There are 578 subjects (chicks)
- `ChickWeight` is already in your workspace

*We will treat the `ChickWeight` data set as the entire population of interest.*

In this exercise we are going take random samples from `ChickWeight` using the `sample()` function.

Recall the syntax to take a random sample and save it to an object:

`sample_name <- sample(x = dataset$variable, size = number)`.

Note: Remember that after we create an object we need to type the object name to view it *(I have already typed this for you in the R Script)*.

*** =instructions
- Take a random sample of 25 from the variable `weight` in the `ChickWeight` data set, assign this to `sample_weight`.
- Take a random sample of 50 from the variable `Diet` in the `ChickWeight` data set, assign this to `sample_diet`.
*** =hint
- For the 1st instruction, follow the format in the lesson but use `ChickWeight$weight` and `25`.

- For the 2nd instruction, follow the format in the lesson but use `ChickWeight$Diet` and `50`.
*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
# Take a random sample of 25 from the variable weight in ChickWeight data set
sample_weight <- 



# Print out the values in sample_weight

sample_weight


# Take a random sample of 50 from the variable Diet in ChickWeight data set
sample_diet <- 



# Print out the values in sample_diet

sample_diet



```

*** =solution
```{r}
# Take a random sample of 25 from the variable weight in ChickWeight data set
sample_weight <- sample(x = ChickWeight$weight, size = 25)

# Print out the values in sample_weight
sample_weight

# Take a random sample of 50 from the variable Diet in ChickWeight data set
sample_diet <- sample(x = ChickWeight$Diet, size = 50)


# Print out the values in sample_diet
sample_diet



```

*** =sct
```{r}
test_object("sample_weight", incorrect_msg = "Make sure you use the `sample()` function on the right side of '<-' with the specified parameters; in this case we're using the `ChickWeight` dataset, `weight` variable, and `25` as the sample size.")

test_output_contains("sample_weight", incorrect_msg = "Do not alter the line of code that says `sample_weight`. This command will print the results to the R output. You can click the 'undo' button to get back to the original code.")



test_object("sample_diet", incorrect_msg = "Make sure you use the `sample()` function on the right side of '<-' with the specified parameters; in this case we're using the `ChickWeight` dataset, `Diet` variable, and `50` as the sample size.")

test_output_contains("sample_diet", incorrect_msg = "Do not alter the line of code that says `sample_diet`. This command will print the results to the R output. You can click the 'undo` button to get back to the original code.")


```



--- type:NormalExercise lang:r xp:100 skills:1 key:7e24794db5
## Population distribution


For this exercise, you are going to find the *population distribution* of the quantitative variable `weight` in the `ChickWeight` data set.

**Remember, we will treat the `ChickWeight` data set as the entire population of interest.**


*** =instructions
- Type `library(mosaic)` to load the *mosaic* package into your workspace (You need this package to use the `favstats()` function).
- Use the `favstats()` function to get summary statistics of `weight` from the `ChickWeight` data set.
- Use the `hist()` function to create a histogram of `weight`.
- Click the 'Submit Answer' button and look at the R output.
*** =hint
- Remember, to find summary statistics use the syntax: `favstats(dataset$variable)`.
- To create a histogram, use the syntax: `hist(dataset$variable)`.
*** =pre_exercise_code
```{r}
library(mosaic)
```

*** =sample_code
```{r}
# Load the mosaic package


# Find summary statistics of chick weights


# Create a histogram of chick weights

```

*** =solution
```{r}
# Load the mosaic package
library(mosaic)

# Find summary statistics of chick weights
favstats(ChickWeight$weight)

# Create a histogram of chick weights
hist(ChickWeight$weight)

```

*** =sct
```{r}

test_student_typed("library(mosaic)",  not_typed_msg =  "Type library(mosaic) to load the mosaic package.")

test_function("favstats", args = "x", incorrect_msg = "Remember, to find summary statistics use format: `favstats(dataset$variable)` with specified parameters.")
test_function("hist", args = "x", incorrect_msg = "To create a histogram, use the format: `hist(dataset$variable)` with the specified parameters.")

success_msg("Nice! Now look at the histogram and summary statistics of chick weights.")

```



--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:595c851a82
## Quick check 1

The *Plots* panel contains the histogram of chick weights from the last exercise. How would you best describe this distribution?
*** =instructions
- Right-skewed
- Left-skewed
- Normal
- Uniform
*** =hint

*** =pre_exercise_code
```{r}
hist(ChickWeight$weight)

```

*** =sct
```{r}
msg_bad <- "That is not correct. The histogram is skewed to the right."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad, msg_bad))
```
--- type:NormalExercise lang:r xp:100 skills:1 key:4face42bf0
## Sample mean and sample distribution

Most of the time we don't know what the *population mean* is, but we can **estimate** this value with a random sample. 

In this exercise, you will:

* create a random sample of 75 from the `weight` variable in the `ChickWeight` data set, and,

* compare the *sample mean* `weight` of chicks to the *population mean* you found in the last exercise.

Recall the syntax to obtain a random sample:

`sample(x = dataset$variable, size = number)`
*** =instructions
- Use `sample()` to take a sample of 75 from the `weight` variable, assign this to `my_sample`.
- Use the `favstats()` function to find summary statistics of `my_sample`.
- Use the `hist()` function on `my_sample`.
- Click 'Submit Answer' and look at the R output.
*** =hint
- To obtain a random sample, use the format `sample(x = dataset$variable, size = number)` with the specified dataset, variable and number.

- `my_sample` is just a vector of 75 chick weights so to obtain summary statistics, all you need to put in the `favstats()` function is the sample name `my_sample`.


*** =pre_exercise_code
```{r}
library(mosaic)


```

*** =sample_code
```{r}
# Use the format sample(x = dataset$variable, size = number) with the specified arguments and the ChickWeight data set
my_sample <- 

# Find summary statistics of my_sample with favstats() 



# Create a histogram of my_sample

```

*** =solution
```{r}
# Use the format sample(x = dataset$variable, size = number) with the specified arguments and the ChickWeight data set

my_sample <- sample(ChickWeight$weight, 75)

# Find summary statistics of my_sample with favstats() 
favstats(my_sample)


# Create a histogram of my_sample
hist(my_sample)

```

*** =sct
```{r}
test_object("my_sample", incorrect_msg = "Make sure you use follow the syntax: `sample(x = dataset$variable, size = number)` on the right side of '<-' with the specified parameters; in this case we're using the `ChickWeight` dataset, `weight` variable, and a sample size of 75.")


test_function("sample", args = c("x", "size"),  incorrect_msg = "To create an object called `my_sample`, type sample(x = ChickWeight$weight, size = 75) on the right side of '<-'.")

test_function("favstats", args = "x", incorrect_msg = "Remember, `my_sample` is just a sample from the `weight` variable. To obtain  summary statistics type `favstats(my_sample)`.")

test_function("hist", args = "x", incorrect_msg = "Remember, `my_sample` is just a sample from the weight variable so to get a histogram type `hist(my_sample)`.")

```




--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:281452fdca
## Quick check 2

Suppose that in `my_sample`, the mean chick weight was 125.65 grams.

How does this compare to the **population mean**?

The summary statistics for the population (the entire data set) are stored in an object called `pop_results`; in your R console type `pop_results` to view.


*** =instructions
- The sample mean is less than the population mean.
- The sample mean is equal to the population mean.
- The sample mean is greater than the population mean.

*** =hint

*** =pre_exercise_code
```{r}
library(mosaic)
pop_results <- favstats(ChickWeight$weight)

```

*** =sct
```{r}
msg_bad <- "That is not correct. Type `pop_results` to view the population mean."
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))
```

--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:8c66be93ac
## Quick check 3

Which function in R allows you to *simultaneously* obtain the mean and standard deviation of quantitative variable?

*** =instructions
- summary()
- head()
- favstats()
- mosaic()
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. The `summary()` function does NOT give you the standard deviation, `head()` prints the first 6 observations, and *mosaic* is a package not a function."
msg_success <- "Exactly!"
test_mc(correct = 3, feedback_msgs = c(msg_bad, msg_bad, msg_success, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:d7a8f825e2
## Confidence interval for a population mean

Assume that the population mean `weight` of chicks is unknown so we obtained a random sample. 

In this exercise, we are going to use R as a calculator to find a 95% confidence interval for the population mean *weight*.

Below is the formula for a confidence interval of a population mean.

$\displaystyle \bar{x} \pm t^* \frac{s}{\sqrt{n}}$

The lower confidence bound is:

$\displaystyle \bar{x} - t^* \frac{s}{\sqrt{n}}$

The upper confidence bound is:

$\displaystyle \bar{x} + t^* \frac{s}{\sqrt{n}}$


*** =instructions
Suppose I sampled 75 of the chicks from `ChickWeight`.  Use R as a calculator to find a 95% confidence interval for the population mean *weight* based on the results below.

* $n=75$
* $\bar{x} = 116.2$
* $s = 67.5$
* $t^* = 1.99$ for n = 75 and 95% confidence 

*Remember, to square root a number use the format: `sqrt(number)`; to multiply two numbers together `5(3)` does NOT work, `5*3` does work.*

*** =hint
Use the formula in the lesson with $\bar{x}$ = 116.2, $s$ = 67.5, $t^*$ = 1.99, and n = 75.

Find 'upper bound' and 'lower bound' seperately.
*** =pre_exercise_code
```{r}


```

*** =sample_code
```{r}
# Lower confidence bound

# Upper confidence bound


```

*** =solution
```{r}
# Lower confidence bound
116.2 - 1.99 * (67.5/sqrt(75))

# Upper confidence bound
116.2 + 1.99 * (67.5/sqrt(75))


```

*** =sct
```{r}
test_output_contains("100.6895", incorrect_msg = "Follow the order of operations from your high school algebra class 'PEMDAS' and remember, when you want to square something in R, use the sqrt() function. For the lower confidence bound, subtract the margin of error from the sample mean.")

test_output_contains("131.7105", incorrect_msg = "Follow the order of operations from your high school algebra class 'PEMDAS' and remember, when you want to square something in R, use the sqrt() function. For the upper confidence bound, add the margin of error to the sample mean.")


success_msg("Good job!")

```


--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:ba8db091cd
## LAST EXERCISE OF LAB 4b

**This is the last exercice of Lab 4b.**

Which of the following commands would be most useful to examine the **shape** of a quantitative variable?

*** =instructions
- barplot(data$variable)
- favstats(data$variable)
- table(data$variable)
- hist(data$variable)
*** =hint
What kind of graph have we been creating in this lab?
*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. To examine the SHAPE of a quantitative variable, we create a histogram."
msg_success <- "Exactly! You finished Lab 4b practice course."
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))

```
