---
title       : Lab 3
description : Summarizing and Visualizing Data

--- type:NormalExercise lang:r xp:100 skills:1 key:fe98ce0ad3
## Using the str() function to reveal variable types

For this lab, we will use a dataset called `CPS85` which contains census data on people in the United States in 1985.

- there are 11 variables in the data set
- there are 534 subjects
- `CPS85` is already in your workspace

If we want to know the *types of variables* in our data set, we can apply the `str()` function with the format:

`str(dataset)`

This will return all of the variable names as well as their type (i.e. numeric, factor, integer). 


*Categorical* variables can be `factor` or `character` variable types.

*Quantitative* variables can be `numeric` or `integer` variable types.



*** =instructions
- Use the `str()` function to find the variable types of the `CPS85` data set.
- Click the 'Submit Answer' Button and take a look at the R output in the console.
*** =hint
Follow the format `str(dataset)` with CPS85 instead of dataset.
*** =pre_exercise_code
```{r}
library(mosaicData)
CPS85$sex <- as.numeric(CPS85$sex)

```

*** =sample_code
```{r}
# Find the variable types of CPS85


```

*** =solution
```{r}
# Find the variable types of CPS85
str(CPS85)


```



*** =sct
```{r}
test_function("str", args = "object", incorrect_msg = "Make sure you follow the format `str(dataset)` with CPS85 instead of dataset!")

success_msg("Great! Now look at the R output and find the variable `sex`. What type of variable does R think this is?") 

```


--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:fdd4d759e9
## Quick check 1
Recall from your previous output that the `sex` variable is coded as `1` and `2`, which represents males and females.

In *reality*, the `sex` variable represents a `______` variable; in *R*, this variable should be defined as a `______` variable.
*** =instructions
- quantitative, numeric
- quantitative, factor
- categorical, numeric
- categorical, factor
*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. Since `sex` only takes on two values it represents a categorical variable in reality and should be coded in R as a factor variable."
msg_success <- "Exactly!"
test_mc(correct = 4, feedback_msgs = c(msg_bad, msg_bad, msg_bad, msg_success))

```





--- type:NormalExercise lang:r xp:100 skills:1 key:53e494b387
## Changing variable types

R assigned the `sex` variable to be *numeric* but really it should be coded as as *factor* because it is a **categorical variable**. 

In this exercise we are going to recode the `sex` variable from a numeric to a factor variable type.


*** =instructions
- The first line of code in your script will recode `sex` from a numeric variable to its' correct variable type, *factor*. **Do not remove or alter this line of code.**
- Use the `str()` function on `CPS85`.
- Click the 'Submit Answer' Button and verify that `sex` was correctly recoded to a factor variable.
*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
CPS85$sex <- as.numeric(CPS85$sex)
```

*** =sample_code
```{r}
# This is the code that will change the sex variable to a factor variable (DO NOT alter it)
CPS85$sex <- factor(CPS85$sex, labels = c("female", "male"))


# Use the str() function on CPS85


```

*** =solution
```{r}
# This is the code that will change the sex variable to a factor variable (DO NOT alter it)
CPS85$sex <- factor(CPS85$sex, labels = c('female', 'male'))

# Use the str() function on CPS85
str(CPS85)
```

*** =sct
```{r}
msg <- "Do not alter the code that changed `sex` into a factor variable. Press the 'undo' button in your script to get back to the original code."

test_student_typed("CPS85$sex <- factor(CPS85$sex, labels = c('female', 'male'))", not_typed_msg = msg)


test_function("str", args = "object", incorrect_msg = "Make sure you follow the format `str(dataset)` with the correct data set!")
```


--- type:NormalExercise lang:r xp:100 skills:1 key:b715998152
## Summarizing quantitative variables

Previously, we used the `summary()` function to summarize data; however, this function does not produce standard deviation.  In order to get all of our standard summary statistics simultaneously, we will use the `favstats()` function from the *mosaic* package.

* First, to access a particular package, we need to load the package with the `library(packagename)` command.
* Next, use the `favstats()` function in the format: `favstats(dataset$quant_var)`.

`CPS85` contains a quantitative variable named `wage` that measures the respondents wage in US dollars per hour.


*** =instructions
- Type `library(mosaic)` to load the *mosaic* package into your workspace.
- Find summary statistics of the `wage` variable in the `CPS85` data set using the `favstats()` function.

*** =hint
Follow the format in the lesson substituting `CPS85` for dataset and `wage` for quant_var.

*** =pre_exercise_code
```{r}
library(mosaicData)
CPS85$sex <- factor(CPS85$sex, labels = c('female', 'male'))
library(mosaic)
```

*** =sample_code
```{r}
# Load the mosaic package 


# Use the favstats() function to explore wage in the CPS85 data set 



```

*** =solution
```{r}
# Load the mosaic package 
library(mosaic)

# Use the favstats() function to explore wage in the CPS85 data set 
favstats(CPS85$wage)



```

*** =sct
```{r}
test_function("favstats", args = "x", index = 1, incorrect_msg = "Follow the format in the lesson with correct dataset and variable!")

```



--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:10bdedd6ea
## Quick check 2

The `favstats()` results of `age` from the last exercise are stored in an object called `age_results`; in your R console type `age_results` to view these.

How old was the oldest respondent in the data set?

*** =instructions
- 28
- 64
- 35
- 44
*** =hint

*** =pre_exercise_code
```{r}
library(mosaic)
library(mosaicData)
age_results <- favstats(CPS85$age)
```

*** =sct
```{r}
msg_bad <- "That is not correct. The oldest person will be the `max` of age in the data set."
msg_success <- "Exactly!"
test_mc(correct = 2, feedback_msgs = c(msg_bad, msg_success, msg_bad, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:b15fdcaef0
## Summarizing quantitative variables by group

What if we wanted to determine if `wage` differs depending on the respondents sex (`sex`)? To do this, we can summarize `wage` separately for the men and women using `favstats()` function with the format:

`favstats(dataset$quant_var ~ dataset$categ_var)`

*Note: the quantitative variable goes on the left hand side of the tilde, and the categorical grouping variable goes on the right hand side.*
*** =instructions
- The `mosaic` package is already loaded in you workspace, so you **do not** need to re-submit the `library(mosaic)` command.
- Use the `favstats()` function to compare the subjects' `wage` by their `sex` in the `CPS85` data set).

*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
library(mosaic)
CPS85$sex <- factor(CPS85$sex, labels = c("female", "male"))

```

*** =sample_code
```{r}
# Compare the average wage among men and women in the CPS85 data set 


```

*** =solution
```{r}
# Compare the average weight among men and women in the CPS85 data set 
favstats(CPS85$wage ~ CPS85$sex)



```

*** =sct
```{r}
test_student_typed("favstats(CPS85$wage ~ CPS85$sex)", not_typed_msg = "Make sure you follow the format in the lesson with correct data set (`CPS85`), quantitative variable (`wage`) and categorical variable (`sex`).")
```


--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:a2c68527ae
## Interpret the results

The results from your previous exercise have been stored in your workspace as `results`.  Submit `results` in the console to review the output. Note that `wage` represents dollars per hour.

There are `_____` males in the study than females; on average, males earned `______` dollars per hour than females.

*** =instructions
- more, more
- more, less
- less, more
- less, less

*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
library(mosaic)
CPS85$sex <- factor(CPS85$sex, labels = c("female", "male"))
results <- favstats(CPS85$wage ~ CPS85$sex)
```

*** =sct
```{r}
msg_bad <- "That is not quite correct. Review the output of `results` carefully."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))
```
--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:fd332565c4
## Quick check 3

When using favstats, the general syntax is:


*** =instructions
- `favstats(categorical_variable ~ categorical_variable)`
- `favstats(quantitative_variable ~ categorical_variable)`
- `favstats(quantitative_variable ~ quantitative_variable)`
- `favstats(categorical_variable ~ quantitative_variable)`

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct."
msg_success <- "Exactly!"
test_mc(correct = 2, feedback_msgs = c(msg_bad, msg_success, msg_bad, msg_bad))

```
--- type:NormalExercise lang:r xp:100 skills:1 key:c7db7fdc55
## Summarizing categorical variables

To summarize categorical variables, we can use the `table()` function to see the frequency of occurences at each level of the categorical variable.

To create a table, use the format:

`table(dataset$categ_var)`

The `CPS85` data set contains a categorical variable named `married` which indicates whether or not the subject is currently married.

*** =instructions
- Summarize the `married` variable in the `CPS85` data set using the `table()` function.
- Click the 'Submit Answer' Button and take a look at the R output in the console.
*** =hint
Use the format from the lesson with `CPS85` instead of dataset, and `married` instead of categ_var.
*** =pre_exercise_code
```{r}
library(mosaicData)

```

*** =sample_code
```{r}
# Type table(dataset$categ_var) with specified data set and variable to get a table summary


```

*** =solution
```{r}
# Type table(dataset$categ_var) with specified data set and variable to get a table summary
table(CPS85$married)

```

*** =sct
```{r}
test_function("table", args = "...", incorrect_msg = "Follow the format in the lesson with correct dataset (`CPS85`) and categ_var (`married`).")

```



--- type:NormalExercise lang:r xp:100 skills:1 key:9fc43281cb
## Contingency tables

A contingency table is used to summarize the relationship betweeen categorical variables. 

In this exercise we will create a contingency table by cross-classifying two variables using the `table()` function with the format:

`table(dataset$var1, dataset$var2)`

Note: The variable listed first (var1) will go along the rows and the variable listed second (var2) will go along the columns.


*** =instructions
- Create a contingency table with variables `married` on the rows and `sex` on the columns from the `CPS85` data set.
*** =hint
Use the format from the lesson with `CPS85` instead of dataset, `married` instead of var1, and `sex` instead of var2.

*** =pre_exercise_code
```{r}
library(mosaicData)
CPS85$sex <- factor(CPS85$sex, labels = c("female", "male"))

```

*** =sample_code
```{r}
# Create contingency table with variables married and sex


```

*** =solution
```{r}
# Create contingency table with variables married and sex
table(CPS85$married, CPS85$sex)


```

*** =sct
```{r}
test_function("table", args = "...", incorrect_msg = "Use the format from the lesson with `CPS85` instead of dataset, `married` instead of var1, and `sex` instead of var2.")
```





--- type:PlainMultipleChoiceExercise lang:r xp:50 skills:1 key:061eb1811c
## Quick check 4

Which function summarizes a ***categorical*** variable in R?

*** =instructions
- table
- type
- favstats
- str
*** =hint
Remember, we have to create a table to summarize a categorical variable.

*** =pre_exercise_code
```{r}

```

*** =sct
```{r}
msg_bad <- "That is not correct. Remember, we have to create a table to summarize a categorical variable."
msg_success <- "Exactly!"
test_mc(correct = 1, feedback_msgs = c(msg_success, msg_bad, msg_bad, msg_bad))

```


--- type:NormalExercise lang:r xp:100 skills:1 key:9a4fea2da0
## Visualizing quantitative data in R
The two main visualizations of quantitative data are *histograms* and *boxplots*.

- use `hist(dataset$quant_var)` to create a histogram

- use `boxplot(dataset$quant_var)` to create a boxplot

*** =instructions
- Create a histogram of the `wage` variable from the `CPS85` data set using the `hist()` function.
- Click the 'Submit Answer' Button and take a look at the R output in the console.
*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
```

*** =sample_code
```{r}
# Create histogram of wage variable with hist() function


```

*** =solution
```{r}
# Create histogram of wage variable with hist() function

hist(CPS85$wage)
```

*** =sct
```{r}
test_function("hist", args = "x", incorrect_msg = "Follow the format: `hist(dataset$variable)` with specified dataset and variable.")
```



--- type:NormalExercise lang:r xp:100 skills:1 key:fca8885b03
## Visualizing quantitative variables by groups

Sometimes we want to compare a quantitative variable across the different groups of a categorical variable. 

To do this, we can use the `boxplot()` function of the format:

`boxplot(dataset$quant_var ~ dataset$categ_var)`

This will produce side by side boxplots.

*** =instructions

Create side-by-side boxplots to compare the `wage` of respondents in the `CPS85` data set by whether they are male or female (`sex`).

*** =hint
Use format in the lesson with `wage` instead of `quant_var`, `CPS85` instead of `dataset`, `sex` instead of `categ_var`.
*** =pre_exercise_code
```{r}
library(mosaicData)
CPS85$sex <- factor(CPS85$sex, labels = c("female", "male"))


```

*** =sample_code
```{r}
# Create side-by-side boxplots where wage is the quantitative variable and sex is categorical


```

*** =solution
```{r}
# Create side-by-side boxplots where wage is the quantitative variable and sex is categorical

boxplot(CPS85$wage ~ CPS85$sex)
```

*** =sct
```{r}
test_student_typed("boxplot(CPS85$wage ~ CPS85$sex)", not_typed_msg = "Remember, `wage` is a quantitative variable and `sex` is a categorical variable. Make sure you type these in the correct order to make the boxplot.")

success_msg("Great job. You finished Lab 3 practice course.")
```


--- type:NormalExercise lang:r xp:100 skills:1 key:b65646dd8a
## Visualizing categorical data in R

One way visualize a categorical variable is with a a *barplot*.

To do this in R, use the format:

`barplot(table(dataset$categ_var))`

*** =instructions
- Create a barplot of the categorical variable `married` from the `CPS85` data set.
- Click the 'Submit Answer' Button and take a look at the R output in the console.
*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
```

*** =sample_code
```{r}
# Create barplot of the married variable in CPS85 data set

```

*** =solution
```{r}
# Create barplot of the married variable in CPS85 dataset
barplot(table(CPS85$married))

```

*** =sct
```{r}
test_function("barplot", args = "height", incorrect_msg = "Make sure you follow the format in the lesson with `CPS85` instead of dataset and `married` instead of categ_var.")

```
--- type:NormalExercise lang:r xp:100 skills:1 key:c6f71ca0bd
## Visualizing relationships between quantitative variables

If we have two quantitative variables, we can visualize their relationship with the `plot()` function. 

The format to make a basic scatterplot is:

`plot(x = dataset$var1 , y = dataset$var2)`

`CPS85` contains a quantitative variable named `educ` which measures the respondents number of years of education.

For this exercise, we are going to visualize the relationship between `educ` and `wage` in the `CPS85` data set. Plot `educ` on the x-axis and `wage` on the y-axis.

*** =instructions
- Use the `plot()` function to create a scatterplot of `educ` and `wage` from the `CPS85` data set.
- After clicking 'Submit Answer', look at the plot in the output. Is there any pattern?
*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
```

*** =sample_code
```{r}
# Use plot() to create scatterplot of education (educ) and wage


```

*** =solution
```{r}
# Use plot() to create scatterplot of education (educ) and wage

plot(x = CPS85$educ, y = CPS85$wage)
```

*** =sct
```{r}
test_function("plot", args = c("x", "y"), incorrect_msg = "Remember to follow the format in the lesson with the specified data set and variables.")

```







--- type:MultipleChoiceExercise lang:r xp:50 skills:1 key:5c4f336256
## LAST EXERCISE OF LAB 3

**This is the last exercice of Lab 3.**

Based on the scatterplot, what do you think is the general relationship between years of education and hourly wage?

*** =instructions
- no association
- a positive association
- a negative association

*** =hint

*** =pre_exercise_code
```{r}
library(mosaicData)
library(mosaic)
CPS85$sex <- factor(CPS85$sex, labels = c("female", "male"))
plot(x = CPS85$educ, y = CPS85$wage)
```

*** =sct
```{r}
msg_bad <- "That is not quite correct. It generally appears that as years of education increases, so does hourly wage.  This indicates a positive association."
msg_success <- "Exactly!  You have completed Lab 3."
test_mc(correct = 2, feedback_msgs = c(msg_bad,msg_success, msg_bad))

```
